---
title: "01 - Decontamination and Sample Filtering"
author: "Nickole Villabona"
date: "2025"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: united
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
# Set working directory to project root so relative paths work when knitting from RStudio
knitr::opts_knit$set(root.dir = normalizePath(".."))

library(vroom)
library(tidyverse)
library(reshape2)
library(funrar)
library(phyloseq)
library(decontam)

`%notin%` <- function(x, y) !(x %in% y)

# ── File paths ───────────────────────────────────────────────────────────────
# Adjust these paths to match your local directory structure
path_raw  <- "data/raw/"
path_proc <- "data/processed/"

# Create processed directory if it doesn't exist
dir.create(path_proc, showWarnings = FALSE, recursive = TRUE)
```

---

## Overview

This script performs decontamination of 16S rRNA amplicon sequencing data using the
`decontam` package, followed by sample-level filtering based on contaminant load.

**Inputs (from `data/raw/`):**
- `OTU_table_nc.tsv` — raw OTU table including negative control samples
- `Orb_Tax_table.txt` — taxonomy table with manual LCA corrections for Neisseriaceae and Orbaceae
- `metadata_table.txt` — sample metadata including negative control flags

**Outputs (to `data/processed/`):**
- `contamdf.prev05.rds` — decontam results object
- `OTUs_AbsAbun.tsv` — filtered OTU table (absolute abundances)
- `muestras_a_eliminar.rds` — vector of samples removed due to high contaminant load
- `metadata_2025.txt` — merged and annotated metadata

---

# 1. Decontamination

> Reference: https://benjjneb.github.io/decontam/vignettes/decontam_intro.html

`decontam` identifies contaminant ASVs by comparing their prevalence in true samples
versus negative controls. We use the prevalence method with a threshold of 0.5, which
flags ASVs that are more prevalent in negative controls than in true samples.

## 1.1 Load data

```{r load-decontam-data}
# Load raw OTU table (includes negative control columns)
OTU_D <- vroom(paste0(path_raw, "OTU_table_nc.tsv"), delim = "\t", col_names = TRUE)

# Load taxonomy table (with manual LCA corrections for Neisseriaceae and Orbaceae)
Tax_D <- vroom(paste0(path_raw, "Orb_Tax_table.txt"), delim = "\t", col_names = TRUE)

# Parse taxonomy string into separate rank columns
Tax_D <- Tax_D %>%
  separate(col = taxonomy,
           into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus"),
           sep  = ";",
           remove = FALSE) %>%
  mutate(across(everything(), ~ gsub(" ", "", .))) %>%
  mutate(across(.cols = c(Domain, Phylum, Class, Order, Family, Genus),
                .fns  = ~ sub("^[a-z]__", "", .))) %>%
  select(-taxonomy)

# Load metadata (includes negative control rows)
Metadata_NV <- vroom(paste0(path_raw, "metadata_table.txt"), delim = "\t", col_names = TRUE)

cat("OTU table dimensions:", nrow(OTU_D), "ASVs x", ncol(OTU_D) - 1, "samples\n")
cat("Taxonomy table dimensions:", nrow(Tax_D), "ASVs\n")
cat("Metadata dimensions:", nrow(Metadata_NV), "samples\n")
```

## 1.2 Build phyloseq object for decontam

```{r build-phyloseq-decontam}
# Prepare metadata
Metadata_NV_ps <- column_to_rownames(Metadata_NV, var = "sample")
META_NV        <- sample_data(Metadata_NV_ps)

# Prepare OTU table
OTU_D_mat <- OTU_D %>%
  column_to_rownames(var = "OTU_ID") %>%
  as.matrix()

# Prepare taxonomy table
Tax_D_mat <- Tax_D %>%
  column_to_rownames(var = "OTU_ID") %>%
  as.matrix()

# Build phyloseq object
OTUps <- otu_table(OTU_D_mat, taxa_are_rows = TRUE)
TAXps <- tax_table(Tax_D_mat)
ps    <- phyloseq(OTUps, TAXps)
ps    <- merge_phyloseq(ps, META_NV)
ps
```

This phyloseq object contains **`r ntaxa(ps)` ASVs** inferred by DADA2 from the V4
region of the 16S rRNA gene, across **`r nsamples(ps)` samples** (including negative controls).

## 1.3 Run decontam (prevalence method)

```{r run-decontam}
# Flag negative control samples
sample_data(ps)$is.neg <- sample_data(ps)$Sample_or_Control == "Control_Sample"

# Default threshold (0.1) — for reference
contamdf.prev <- isContaminant(ps, method = "prevalence", neg = "is.neg")
cat("Contaminants at threshold 0.1:", sum(contamdf.prev$contaminant), "\n")

# Stringent threshold (0.5): flags ASVs more prevalent in controls than true samples
# This is the threshold used in the final analysis
contamdf.prev05 <- isContaminant(ps, method = "prevalence", neg = "is.neg", threshold = 0.5)
cat("Contaminants at threshold 0.5:", sum(contamdf.prev05$contaminant), "\n")
```

## 1.4 Visualize contaminants

```{r visualize-contaminants, fig.width=7, fig.height=5}
# Build presence-absence phyloseq objects for visualization
ps.pa     <- transform_sample_counts(ps, function(abund) 1 * (abund > 0))
ps.pa.neg <- prune_samples(sample_data(ps.pa)$Sample_or_Control == "Control_Sample", ps.pa)
ps.pa.pos <- prune_samples(sample_data(ps.pa)$Sample_or_Control == "True_Sample",    ps.pa)

# Prevalence data frame
df.pa <- data.frame(
  pa.pos      = taxa_sums(ps.pa.pos),
  pa.neg      = taxa_sums(ps.pa.neg),
  contaminant = contamdf.prev$contaminant
)

ggplot(data = df.pa, aes(x = pa.neg, y = pa.pos, color = contaminant)) +
  geom_point(alpha = 0.6) +
  xlab("Prevalence (Negative Controls)") +
  ylab("Prevalence (True Samples)") +
  scale_color_manual(values = c("FALSE" = "steelblue", "TRUE" = "coral2")) +
  theme_minimal()
```

## 1.5 Remove contaminants and export decontam results

```{r remove-contaminants}
# Remove contaminant ASVs (threshold = 0.5)
ps.noncontam <- prune_taxa(!contamdf.prev05$contaminant, ps)
cat("ASVs after decontam:", ntaxa(ps.noncontam), "\n")
cat("ASVs removed:", ntaxa(ps) - ntaxa(ps.noncontam), "\n")

# Save decontam results for downstream use
saveRDS(contamdf.prev05, file = paste0(path_proc, "contamdf.prev05.rds"))
```

---

# 2. Sample-Level Filtering

Samples where >25% of ASVs are flagged as contaminants are excluded from further analysis.
This threshold was chosen following best practices for insect microbiome studies
(Williamson et al. 2025).

## 2.1 Calculate contaminant proportion per sample

```{r calculate-contam-proportion}
# Convert to relative abundances for proportion calculation
OTUs_rel <- as.data.frame(make_relative(t(OTU_D_mat))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column(var = "OTU_ID") %>%
  melt(id.vars = "OTU_ID", variable.name = "SampleID", value.name = "RelAbun")

# Join with contaminant flags
decon_df  <- contamdf.prev05 %>% rownames_to_column(var = "OTU_ID")
OTU_decont <- OTUs_rel %>% left_join(decon_df, by = "OTU_ID")

# Count contaminant vs. non-contaminant ASVs per sample
summary_df <- OTU_decont %>%
  filter(RelAbun > 0) %>%
  group_by(SampleID) %>%
  dplyr::count(contaminant) %>%
  reshape2::dcast(SampleID ~ contaminant, value.var = "n") %>%
  setNames(c("SampleID", "N_Clean", "N_Contam")) %>%
  mutate(
    N_Clean         = replace_na(N_Clean,  0),
    N_Contam        = replace_na(N_Contam, 0),
    PropContaminant = N_Contam / (N_Clean + N_Contam)
  ) %>%
  arrange(desc(PropContaminant))
```

## 2.2 Visualize contaminant proportion per sample

```{r plot-contam-proportion, fig.width=12, fig.height=5}
# Join proportion back to full table for plotting
OTU_decont2 <- OTU_decont %>%
  left_join(summary_df[, c("SampleID", "PropContaminant")], by = "SampleID")

ggplot(OTU_decont2,
       aes(x = reorder(SampleID, PropContaminant), y = RelAbun, fill = contaminant)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_hline(yintercept = 0.25, color = "coral2", linetype = "dashed") +
  labs(y = "Relative Abundance", x = "Sample",
       fill = "Contaminant",
       title = "Contaminant proportion per sample (dashed line = 25% threshold)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 6))
```

## 2.3 Remove high-contaminant samples and contaminant ASVs

```{r filter-samples}
# Samples with >25% contaminant ASVs are removed (paper: "more than 25%")
muestras_a_eliminar <- summary_df %>%
  filter(PropContaminant > 0.25) %>%
  pull(SampleID) %>%
  as.character()

cat("Samples removed due to high contaminant load (>25%):\n")
print(muestras_a_eliminar)
cat("Total samples removed:", length(muestras_a_eliminar), "\n")

# Save for downstream use in 03_qPCR script
saveRDS(muestras_a_eliminar, file = paste0(path_proc, "muestras_a_eliminar.rds"))

# Remove flagged samples and all remaining contaminant ASVs
OTUs_filtered <- OTU_decont %>%
  filter(SampleID %notin% muestras_a_eliminar,
         contaminant == FALSE)

# Pivot back to wide format (OTU_ID x Sample)
OTUs_filtered_wide <- reshape2::dcast(OTUs_filtered, OTU_ID ~ SampleID,
                                       value.var = "RelAbun") %>%
  column_to_rownames(var = "OTU_ID") %>%
  mutate_all(as.numeric)

cat("Samples remaining:", ncol(OTUs_filtered_wide), "\n")
cat("ASVs remaining:", nrow(OTUs_filtered_wide), "\n")
```

## 2.4 Recover absolute abundances for filtered ASVs and samples

We use the raw (pre-decontam) OTU table to recover absolute abundances for the
retained ASVs and samples, as the decontam step used relative abundances internally.

```{r recover-absolute}
samples_to_keep <- colnames(OTUs_filtered_wide)
asvs_to_keep    <- rownames(OTUs_filtered_wide)

# Subset raw table to retained ASVs and samples
OTUs_AbsAbun <- OTU_D %>%
  select(OTU_ID, any_of(samples_to_keep)) %>%
  filter(OTU_ID %in% asvs_to_keep) %>%
  column_to_rownames(var = "OTU_ID")

cat("Final OTU table dimensions:",
    nrow(OTUs_AbsAbun), "ASVs x", ncol(OTUs_AbsAbun), "samples\n")

# Export for downstream use
write.table(
  cbind("ASV_ID" = rownames(OTUs_AbsAbun), OTUs_AbsAbun),
  file      = paste0(path_proc, "OTUs_AbsAbun.tsv"),
  sep       = "\t",
  row.names = FALSE,
  quote     = FALSE
)

cat("Exported: OTUs_AbsAbun.tsv\n")
```

---

# 3. Metadata Preparation

```{r load-metadata}
# Extended metadata from Claus et al. collection
BigMetadata <- vroom(paste0(path_raw, "MappingFile_Meliponini_Claus.txt"),
                     delim = "\t", col_names = TRUE)

# Merge metadata tables
metadata <- merge(Metadata_NV, BigMetadata,
                  by.x = "CollectionID", by.y = "ID_collection") %>%
  select(-matches("Genus\\.x|Species\\.x")) %>%
  rename_with(~ "Species", matches("Species\\.y")) %>%
  rename_with(~ "Genus",   matches("Genus\\.y"))

# Add geographic groupings
metadata <- metadata %>%
  mutate(
    Continent = case_when(
      Country %in% c("Peru", "Brazil", "Colombia", "Bolivia",
                      "Venezuela", "Ecuador")                        ~ "South_America",
      Country %in% c("Costa Rica", "Guatemala", "Monserrat",
                      "Panama", "Mexico")                            ~ "N_C_America",
      Country %in% c("Australia")                                    ~ "Australia",
      Country %in% c("Cameroon", "Madagascar", "South Africa",
                      "Uganda")                                      ~ "Africa",
      Country %in% c("India", "Indonesia", "Malaysia")              ~ "Asia"
    ),
    Region = case_when(
      Continent %in% c("South_America", "N_C_America") ~ "Neotropics",
      Continent %in% c("Africa")                       ~ "AfroTropics",
      Continent %in% c("Asia", "Australia")            ~ "Indo-Malay-Australasia"
    )
  )

colnames(metadata) <- c("CollectionID", "sample", "Box", "PrepDate", "BodySize",
                         "ColonySize_L", "ColonySize_U", "ColonySize",
                         "Sample_or_Control", "Genus", "Species", "Number_Bees",
                         "Country", "Area", "Preservation", "Continent", "Region")

# Export merged metadata
write.table(metadata,
            file      = paste0(path_proc, "metadata_2025.txt"),
            sep       = "\t",
            row.names = FALSE,
            quote     = FALSE)

cat("Exported: metadata_2025.txt\n")
cat("Metadata dimensions:", nrow(metadata), "samples x", ncol(metadata), "variables\n")
```

---

# 4. Summary of Outputs

```{r summary}
cat("============================================\n")
cat("Files exported to data/processed/:\n")
cat("  - contamdf.prev05.rds\n")
cat("  - muestras_a_eliminar.rds\n")
cat("  - OTUs_AbsAbun.tsv\n")
cat("  - metadata_2025.txt\n")
cat("============================================\n")
cat("These files are required inputs for:\n")
cat("  → 02_Microbiome_Analysis.Rmd\n")
cat("  → 03_qPCR_and_AbsAbundance.Rmd\n")
```

---

# Session Info

```{r session-info}
sessionInfo()
```
