---
title: "02 - Microbiome Analysis and Phylogenetic Comparative Methods"
author: "Nickole Villabona"
date: "2025"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
# Set working directory to project root so relative paths work when knitting from RStudio
knitr::opts_knit$set(root.dir = normalizePath(".."))

library(vroom)
library(tidyverse)
library(reshape2)
library(funrar)
library(phyloseq)
library(microViz)
library(vegan)
library(ggpubr)
library(RColorBrewer)
library(gridExtra)
library(ape)
library(phytools)
library(nlme)
library(corHMM)

`%notin%` <- function(x, y) !(x %in% y)

# ── File paths ───────────────────────────────────────────────────────────────
path_raw  <- "data/raw/"
path_proc <- "data/processed/"

dir.create("figures/raw", showWarnings = FALSE, recursive = TRUE)
```

---

## Overview

This script performs microbiome community analysis and phylogenetic comparative methods
to test associations between gut symbiont dynamics and host colony size.

**Inputs (from `data/processed/`):**
- `OTUs_AbsAbun.tsv` — filtered OTU table (absolute abundances)
- `metadata_2025.txt` — merged and annotated metadata

**Inputs (from `data/raw/`):**
- `Orb_Tax_table.txt` — taxonomy table with manual LCA corrections
- `MappingFile_Meliponini_Claus.txt` — extended collection metadata
- `meliponini5genes.nex` — stingless bee phylogeny (Rasmussen & Cameron 2010)

**Outputs (to `data/processed/`):**
- `phyloseqOBJ.rds` — rarefied and filtered phyloseq object
- `alpha_table.rds` — alpha diversity table with metadata
- `metadata_symbionts.rds` — symbiont prevalence merged with metadata
- `pruned_tree.rds` — ultrametric tree pruned to study species
- `discrete_Sng.rds` — binary Snodgrassella presence/absence per voucher
- `discrete_Gill.rds` — binary Gilliamella presence/absence per voucher
- `discrete_Lacto.rds` — binary Lactobacillus presence/absence per voucher
- `discrete_Comm.rds` — binary Commensalibacter presence/absence per voucher
- `unique_all_avg.rds` — species-averaged symbiont data for PGLS

---

# 1. Build Phyloseq Object

## 1.1 Load data

```{r load-data}
# Load filtered OTU table (absolute abundances, output from script 01)
OTU_Table <- vroom(paste0(path_proc, "OTUs_AbsAbun.tsv"),
                   delim = "\t", col_names = TRUE) %>%
  as.data.frame() %>%
  column_to_rownames(var = "ASV_ID")

# Load taxonomy (with manual LCA corrections for Neisseriaceae and Orbaceae)
Tax <- vroom(paste0(path_raw, "Orb_Tax_table.txt"),
             delim = "\t", col_names = TRUE) %>%
  separate(col = taxonomy,
           into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus"),
           sep  = ";", remove = FALSE) %>%
  mutate(across(everything(), ~ gsub(" ", "", .))) %>%
  mutate(across(.cols = c(Domain, Phylum, Class, Order, Family, Genus),
                .fns  = ~ sub("^[a-z]__", "", .))) %>%
  select(-taxonomy) %>%
  column_to_rownames(var = "OTU_ID") %>%
  as.matrix()

# Load metadata (output from script 01)
metadata <- vroom(paste0(path_proc, "metadata_2025.txt"),
                  delim = "\t", col_names = TRUE) %>%
  as.data.frame()

cat("OTU table:", nrow(OTU_Table), "ASVs x", ncol(OTU_Table), "samples\n")
cat("Taxonomy:", nrow(Tax), "ASVs\n")
cat("Metadata:", nrow(metadata), "samples\n")
```

## 1.2 Build phyloseq object

```{r build-phyloseq}
# Rename species/genus columns to avoid conflicts with taxonomy slot
metadata_ps <- metadata %>%
  rename(Bee_Species = Species, Bee_Genus = Genus) %>%
  column_to_rownames(var = "sample") %>%
  sample_data()

OTUs_ps <- otu_table(OTU_Table, taxa_are_rows = TRUE)
TAXps   <- tax_table(Tax)
ps0     <- phyloseq(OTUs_ps, TAXps)

# Merge and fix taxonomy
# microViz cannot handle empty taxonomic levels; upper-level classification is propagated
phyloseqOBJ_Raw <- merge_phyloseq(ps0, metadata_ps) %>%
  tax_fix(unknowns = c("uncultured"))

phyloseqOBJ_Raw
```

## 1.3 Rarefaction curves

```{r rarefaction-curves, fig.width=12, fig.height=8}
otu_mat      <- as(otu_table(phyloseqOBJ_Raw), "matrix")
if (taxa_are_rows(phyloseqOBJ_Raw)) otu_mat <- t(otu_mat)

rare_list    <- rarecurve(otu_mat, step = 100,
                          sample = min(rowSums(otu_mat)), label = FALSE)
sample_names <- rownames(otu_mat)

rare_df <- map_dfr(seq_along(rare_list), function(i) {
  tibble(
    Sample_ID         = sample_names[i],
    Depth             = attr(rare_list[[i]], "Subsample"),
    Rarefied_Richness = rare_list[[i]]
  )
}) %>%
  left_join(metadata %>% rename(Sample_ID = sample), by = "Sample_ID")

ggplot(rare_df, aes(x = Depth, y = Rarefied_Richness,
                    group = Sample_ID, color = Sample_ID)) +
  geom_line(alpha = 0.8) +
  facet_wrap(~ Genus) +
  geom_vline(xintercept = 2500, color = "darkorange", linetype = "dashed") +
  coord_cartesian(xlim = c(0, 4000)) +
  labs(x = "Number of reads", y = "Observed richness",
       title = "Rarefaction curves per sample (dashed line = 2,500 reads)") +
  theme_minimal() +
  theme(legend.position = "none")

ggsave("figures/raw/rarefaction_curves.pdf", width = 12, height = 8)
```

## 1.4 Rarefy to 2,500 reads

```{r rarefy}
# Distribution of library sizes
sample_sums_df <- data.frame(
  SampleID = sample_names(phyloseqOBJ_Raw),
  Reads    = sample_sums(phyloseqOBJ_Raw)
)

ggplot(sample_sums_df, aes(x = Reads)) +
  geom_histogram(binwidth = 400, fill = "cornflowerblue", color = "white") +
  geom_vline(xintercept = 2500, color = "coral2", linetype = "dashed") +
  labs(title = "Distribution of library sizes",
       x = "Reads per sample", y = "Number of samples") +
  theme_minimal()

set.seed(123)
# Rarefy to 2,500 reads per sample (seed = 123 for reproducibility)
phyloseqOBJ_Rarefied <- rarefy_even_depth(
  phyloseqOBJ_Raw,
  sample.size = 2500,
  rngseed     = 123,
  replace     = FALSE,
  trimOTUs    = TRUE,
  verbose     = TRUE
)
```

## 1.5 Prevalence filter and taxonomy cleanup

```{r prevalence-filter}
# Remove very rare ASVs and non-bacterial taxa
phyloseqOBJ <- phyloseqOBJ_Rarefied %>%
  tax_filter(
    min_prevalence           = 0,
    prev_detection_threshold = 0,
    min_total_abundance      = 0,
    min_sample_abundance     = 10,
    tax_level                = NA
  ) %>%
  phyloseq::subset_taxa(
    Domain != "Eukaryota" &
    !is.na(Phylum) &
    Phylum != "Unassigned"
  )

cat("Final dataset:", ntaxa(phyloseqOBJ), "ASVs x",
    nsamples(phyloseqOBJ), "samples\n")

# Save phyloseq object for downstream use
saveRDS(phyloseqOBJ, file = paste0(path_proc, "phyloseqOBJ.rds"))
```

---

# 2. Alpha Diversity

Alpha diversity (Shannon index) is estimated by iteratively rarefying the unfiltered
absolute-abundance table 100 times at a depth of 1,300 reads, then averaging across
iterations to reduce stochastic variation.

```{r alpha-diversity}
n_iterations   <- 100
sequence_depth <- 1300
alpha_results  <- vector("list", n_iterations)

set.seed(711)

for (i in seq_len(n_iterations)) {
  cat("Iteration", i, "of", n_iterations, "\r")
  rar_otu_temp       <- as.data.frame(vegan::rrarefy(OTU_Table, sequence_depth))
  alpha_results[[i]] <- as.data.frame(vegan::diversity(t(rar_otu_temp)))
}

alpha_diversity_table <- data.frame(
  Shannon = rowMeans(do.call(cbind, alpha_results))
)

# Merge with metadata
alpha.table <- merge(alpha_diversity_table, metadata,
                     by.x = "row.names", by.y = "sample")

alpha.table$ColonySize <- factor(alpha.table$ColonySize,
                                  levels = c("Small", "Medium", "Large",
                                             "XLarge", "Enormous"))
alpha.table$Continent  <- factor(alpha.table$Continent,
                                  levels = c("South_America", "N_C_America",
                                             "Australia", "Africa", "Asia"))
alpha.table$Region     <- factor(alpha.table$Region,
                                  levels = c("Neotropics",
                                             "Indo-Malay-Australasia",
                                             "AfroTropics"))

saveRDS(alpha.table, file = paste0(path_proc, "alpha_table.rds"))
cat("\nAlpha diversity table saved.\n")
```

## 2.1 Alpha diversity plots

```{r alpha-plots, fig.width=12, fig.height=4}
A <- ggplot(alpha.table, aes(ColonySize, Shannon)) +
  geom_violin(aes(color = ColonySize)) +
  geom_boxplot(width = 0.1, color = "grey", alpha = 0.2) +
  geom_jitter(alpha = 0.5) +
  theme_classic() +
  theme(text = element_text(size = 14), legend.position = "none")

B <- ggplot(alpha.table, aes(Continent, Shannon)) +
  geom_violin(aes(color = Continent)) +
  geom_boxplot(width = 0.1, color = "grey", alpha = 0.2) +
  geom_jitter(alpha = 0.5) +
  theme_classic() +
  theme(text = element_text(size = 14), legend.position = "none")

C <- ggplot(alpha.table, aes(Region, Shannon)) +
  geom_violin(aes(color = Region)) +
  geom_boxplot(width = 0.1, color = "grey", alpha = 0.2) +
  geom_jitter(alpha = 0.5) +
  theme_classic() +
  theme(text = element_text(size = 14), legend.position = "none")

ggarrange(A, B, C, nrow = 1)
ggsave("figures/raw/alpha_diversity.pdf", width = 12, height = 4)
```

---

# 3. Community Composition

## 3.1 PCA (CLR-transformed, family level)

PCA uses the centred log-ratio (CLR) transformation, which is recommended for
compositional sequencing data.

```{r pca-plot, fig.width=8, fig.height=6}
pca_plot <- phyloseqOBJ %>%
  tax_fix(unknowns = c("Incertae_Sedis", "uncultured", "Unknown")) %>%
  tax_transform("clr", rank = "Genus") %>%
  ord_calc(method = "PCA") %>%
  ord_plot(color = "Region", plot_taxa = 1:8,
           alpha = 0.75,
           auto_caption = 10,
           tax_lab_style = tax_lab_style(size = 2.5, alpha = 0.75))

pca_plot
ggsave("figures/raw/PCA_region.pdf", width = 8, height = 6)
```

## 3.1b PCA loadings — quantify genus contributions to compositional variation

```{r pca-loadings}
# Save the ordination object separately to extract loadings
pca_ord <- phyloseqOBJ %>%
  tax_fix(unknowns = c("Incertae_Sedis", "uncultured", "Unknown")) %>%
  tax_transform("clr", rank = "Genus") %>%
  ord_calc(method = "PCA")

# Extract species scores directly from vegan rda internal structure
# (avoids scores() dispatch issues across microViz/vegan versions)
rda_obj      <- ord_get(pca_ord)
eig          <- rda_obj$CA$eig          # eigenvalues
v            <- rda_obj$CA$v            # raw eigenvectors (taxa × axes)
# Species scaling = eigenvectors × sqrt(eigenvalue)
loadings_mat <- sweep(v[, 1:2, drop = FALSE], 2, sqrt(eig[1:2]), "*")
colnames(loadings_mat) <- c("PC1", "PC2")

# % contribution of each genus to variance along PC1 and PC2
# (squared loading / sum of squared loadings) × 100
loading_table <- as.data.frame(loadings_mat) %>%
  rownames_to_column("Genus") %>%
  mutate(
    PC1_loading     = round(PC1, 4),
    PC1_contrib_pct = round(PC1^2 / sum(PC1^2) * 100, 2),
    PC2_loading     = round(PC2, 4),
    PC2_contrib_pct = round(PC2^2 / sum(PC2^2) * 100, 2)
  ) %>%
  select(Genus, PC1_loading, PC1_contrib_pct, PC2_loading, PC2_contrib_pct) %>%
  arrange(desc(abs(PC1_loading)))

cat("=== PCA loadings ranked by |PC1| ===\n")
print(head(loading_table, 20))

cat("\nTop contributors to PC1 (Gilliamella + Snodgrassella):\n")
print(head(loading_table, 2))
cat(sprintf("\nCombined PC1 contribution: %.1f%%\n",
            sum(head(loading_table, 2)$PC1_contrib_pct)))
```

## 3.2 PCoA (Bray-Curtis, genus level)

```{r pcoa-plot, fig.width=7, fig.height=6}
PCoA_region <- phyloseqOBJ %>%
  tax_agg(rank = "Genus") %>%
  dist_calc(dist = "bray") %>%
  ord_calc(method = "PCoA") %>%
  ord_plot(alpha = 0.6, size = 2, color = "Region") +
  theme_classic(12) +
  coord_fixed(0.7) +
  stat_ellipse(aes(color = Region)) +
  scale_color_brewer(palette = "Set1")

PCoA_region
ggsave("figures/raw/PCoA_region.pdf", width = 7, height = 6)
```

## 3.3 Composition barplot

```{r composition-plots, fig.width=12, fig.height=6}
# Aggregate to genus level and compute relative abundances
ps_genus_rel <- tax_glom(phyloseqOBJ, taxrank = "Genus") %>%
  transform_sample_counts(function(x) x / sum(x))

# Melt to long format (Taxon = bacterial genus; Bee_Genus, Region from sample_data)
df_comp <- psmelt(ps_genus_rel) %>%
  rename(Taxon = Genus)

# Identify top 13 genera by mean relative abundance across all samples
top13 <- df_comp %>%
  group_by(Taxon) %>%
  summarise(mean_abun = mean(Abundance), .groups = "drop") %>%
  arrange(desc(mean_abun)) %>%
  slice_head(n = 13) %>%
  pull(Taxon)

# Lump remaining taxa into "Other"
df_comp <- df_comp %>%
  mutate(Taxon_plot = factor(if_else(Taxon %in% top13, Taxon, "Other"),
                             levels = c(top13, "Other")))

# Order samples by bee genus, then sample ID within each genus
sample_order <- df_comp %>%
  select(Sample, Bee_Genus) %>%
  distinct() %>%
  arrange(Bee_Genus, Sample) %>%
  pull(Sample)

df_comp$Sample <- factor(df_comp$Sample, levels = sample_order)

# Bacterial taxon color palette (Option B: Set2 + Set1)
# ── Option A: Tableau-inspired ────────────────────────────────────────────
# palette_colors <- c("#4E79A7", "#F28E2B", "#E15759", "#76B7B2", "#59A14F",
#                     "#EDC948", "#B07AA1", "#FF9DA7", "#9C755F", "#BAB0AC",
#                     "#D37295", "#FABFD2", "#8CD17D", "grey70")
# ── Option B: Set2 + Set1 (RColorBrewer) ──────────────────────────────────
palette_colors <- c(brewer.pal(8, "Set2"), brewer.pal(5, "Set1"), "grey70")
# ── Option C: vivid qualitative ───────────────────────────────────────────
# palette_colors <- c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00",
#                     "#A65628", "#F781BF", "#999999", "#66C2A5", "#FC8D62",
#                     "#8DA0CB", "#E78AC3", "#A6D854", "grey60")

taxon_colors <- setNames(palette_colors, c(top13, "Other"))

# Annotation data: one row per sample, same factor order as main plot
annot_df <- df_comp %>%
  select(Sample, Bee_Genus, Region) %>%
  distinct() %>%
  mutate(Sample = factor(Sample, levels = levels(df_comp$Sample)))

# Region colors — grayscale
region_cols <- c("Neotropics"             = "grey80",
                 "AfroTropics"            = "grey45",
                 "Indo-Malay-Australasia" = "grey15")

# ── Main barplot ──────────────────────────────────────────────────────────
p_main <- ggplot(df_comp, aes(x = Sample, y = Abundance, fill = Taxon_plot)) +
  geom_bar(stat = "identity", width = 0.85) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), expand = c(0, 0)) +
  scale_fill_manual(values = taxon_colors) +
  labs(x = NULL, y = "Relative Abundance", fill = "Genus") +
  theme_bw(base_size = 10) +
  theme(axis.text.x   = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 4),
        panel.grid    = element_blank(),
        panel.border  = element_blank(),
        axis.line.y   = element_line(colour = "black"),
        legend.position = "right",
        legend.text   = element_text(face = "italic", size = 8),
        plot.margin   = margin(5, 5, 2, 5))

# ── Region annotation strip — grayscale ──────────────────────────────────
p_region <- ggplot(annot_df, aes(x = Sample, y = 1, fill = Region)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_manual(values = region_cols, name = "Region") +
  scale_y_continuous(expand = c(0, 0)) +
  labs(y = "Region") +
  theme_void() +
  theme(legend.position  = "right",
        legend.key.size  = unit(0.5, "cm"),
        legend.text      = element_text(size = 9),
        legend.title     = element_text(size = 9, face = "bold"),
        axis.title.y     = element_text(size = 7, angle = 0, vjust = 0.5, hjust = 1),
        plot.margin      = margin(0, 5, 2, 5))

# ── Combine ───────────────────────────────────────────────────────────────
legend_region  <- cowplot::get_legend(p_region)
p_region_noleg <- p_region + theme(legend.position = "none")

left_col <- cowplot::plot_grid(
  p_main + theme(legend.position = "none"),
  p_region_noleg,
  ncol = 1, align = "v", axis = "lr",
  rel_heights = c(8, 0.4)
)

right_col <- cowplot::plot_grid(
  cowplot::get_legend(p_main),
  legend_region,
  ncol = 1, rel_heights = c(3, 1)
)

cowplot::plot_grid(left_col, right_col, ncol = 2, rel_widths = c(10, 1.5))

ggsave("figures/raw/composition_barplot.pdf", width = 12, height = 7)
```

## 3.3b Composition barplot (microViz)

```{r composition-barplot-microviz, fig.width=14, fig.height=5}
# Order samples by bee genus then sample ID — to match original circular Fig. 1
ordered_samples <- data.frame(sample_data(phyloseqOBJ), stringsAsFactors = FALSE) %>%
  rownames_to_column("SampleID") %>%
  arrange(Bee_Genus, SampleID) %>%
  pull(SampleID)

p_mv <- phyloseqOBJ %>%
  comp_barplot(
    tax_level   = "Genus",
    n_taxa      = 10,
    other_name  = "Other",
    sample_order = "default",
    taxon_renamer = function(x) stringr::str_remove(x, " [ae]t rel."),
    palette     = distinct_palette(n = 10, add = "grey90"),
    merge_other = FALSE,
    bar_outline_colour = NA
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.01))) +
  labs(x = NULL, y = "Relative Abundance") +
  theme_bw(base_size = 10) +
  theme(axis.text.x    = element_blank(),
        axis.ticks.x   = element_blank(),
        panel.grid     = element_blank(),
        legend.text    = element_text(face = "italic", size = 8),
        plot.margin    = margin(5, 5, 0, 5))

# Extract actual sample order from the built ggplot (discrete scale limits = sample names in order)
actual_order <- ggplot_build(p_mv)$layout$panel_params[[1]]$x$limits

# Region annotation strip — aligned to actual barplot sample order
annot_mv <- data.frame(sample_data(phyloseqOBJ), stringsAsFactors = FALSE) %>%
  rownames_to_column("SampleID") %>%
  filter(SampleID %in% actual_order) %>%
  mutate(SampleID = factor(SampleID, levels = actual_order))

region_cols <- c("Neotropics"             = "grey80",
                 "AfroTropics"            = "grey45",
                 "Indo-Malay-Australasia" = "grey15")

p_region_mv <- ggplot(annot_mv, aes(x = SampleID, y = 1, fill = Region)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_manual(values = region_cols, name = "Region") +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = NULL, y = "Region") +
  theme_void() +
  theme(legend.position  = "right",
        legend.key.size  = unit(0.4, "cm"),
        legend.text      = element_text(size = 8),
        legend.title     = element_text(size = 8, face = "bold"),
        axis.title.y     = element_text(size = 7, angle = 0, vjust = 0.5, hjust = 1),
        axis.text.x      = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 3),
        plot.margin      = margin(0, 5, 2, 5))

left_col <- cowplot::plot_grid(
  p_mv + theme(legend.position = "none"),
  p_region_mv + theme(legend.position = "none"),
  ncol = 1, align = "v", axis = "lr",
  rel_heights = c(8, 0.5)
)

right_col <- cowplot::plot_grid(
  cowplot::get_legend(p_mv),
  cowplot::get_legend(p_region_mv),
  ncol = 1, rel_heights = c(3, 1)
)

cowplot::plot_grid(left_col, right_col, ncol = 2, rel_widths = c(10, 1.5))

ggsave("figures/raw/composition_barplot_microviz.pdf", width = 14, height = 5)
```

## 3.3b-ii Composition barplot — ordered by Bee_Genus

```{r composition-barplot-beegenusor, fig.width=14, fig.height=5}
ordered_genus <- data.frame(sample_data(phyloseqOBJ), stringsAsFactors = FALSE) %>%
  rownames_to_column("SampleID") %>%
  arrange(Bee_Genus, SampleID) %>%
  pull(SampleID)

p_mv_genus <- phyloseqOBJ %>%
  comp_barplot(
    tax_level   = "Genus",
    n_taxa      = 10,
    other_name  = "Other",
    sample_order = ordered_genus,
    taxon_renamer = function(x) stringr::str_remove(x, " [ae]t rel."),
    palette     = distinct_palette(n = 10, add = "grey90"),
    merge_other = FALSE,
    bar_outline_colour = NA
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.01))) +
  labs(x = NULL, y = "Relative Abundance") +
  theme_bw(base_size = 10) +
  theme(axis.text.x    = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 5),
        axis.ticks.x   = element_line(linewidth = 0.3),
        panel.grid     = element_blank(),
        legend.text    = element_text(face = "italic", size = 8),
        plot.margin    = margin(5, 5, 0, 5))

annot_genus <- data.frame(sample_data(phyloseqOBJ), stringsAsFactors = FALSE) %>%
  rownames_to_column("SampleID") %>%
  filter(SampleID %in% ordered_genus) %>%
  mutate(SampleID = factor(SampleID, levels = ordered_genus))

region_cols <- c("Neotropics"             = "grey80",
                 "AfroTropics"            = "grey45",
                 "Indo-Malay-Australasia" = "grey15")

p_region_genus <- ggplot(annot_genus, aes(x = SampleID, y = 1, fill = Region)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_manual(values = region_cols, name = "Region") +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = NULL, y = "Region") +
  theme_void() +
  theme(legend.position  = "right",
        legend.key.size  = unit(0.4, "cm"),
        legend.text      = element_text(size = 8),
        legend.title     = element_text(size = 8, face = "bold"),
        axis.title.y     = element_text(size = 7, angle = 0, vjust = 0.5, hjust = 1),
        plot.margin      = margin(0, 5, 2, 5))

left_col_g <- cowplot::plot_grid(
  p_mv_genus + theme(legend.position = "none"),
  p_region_genus + theme(legend.position = "none"),
  ncol = 1, align = "v", axis = "lr",
  rel_heights = c(8, 0.5)
)

right_col_g <- cowplot::plot_grid(
  cowplot::get_legend(p_mv_genus),
  cowplot::get_legend(p_region_genus),
  ncol = 1, rel_heights = c(3, 1)
)

cowplot::plot_grid(left_col_g, right_col_g, ncol = 2, rel_widths = c(10, 1.5))

ggsave("figures/raw/composition_barplot_beegenusor.pdf", width = 14, height = 5)
```

## 3.3c Composition barplot — unrarefied, unfiltered (for comparison)

```{r composition-barplot-unrarefied, fig.width=14, fig.height=5}
# Uses phyloseqOBJ_Raw: after tax_fix but before rarefaction and tax_filter
# Transforms to relative abundance so samples are comparable despite unequal depths
ps_raw_rel <- transform_sample_counts(phyloseqOBJ_Raw, function(x) x / sum(x))

p_mv_raw <- ps_raw_rel %>%
  comp_barplot(
    tax_level   = "Genus",
    n_taxa      = 10,
    other_name  = "Other",
    sample_order = "default",
    taxon_renamer = function(x) stringr::str_remove(x, " [ae]t rel."),
    palette     = distinct_palette(n = 10, add = "grey90"),
    merge_other = FALSE,
    bar_outline_colour = NA
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.01))) +
  labs(x = NULL, y = "Relative Abundance",
       title = "Unrarefied, unfiltered (phyloseqOBJ_Raw)") +
  theme_bw(base_size = 10) +
  theme(axis.text.x    = element_blank(),
        axis.ticks.x   = element_blank(),
        panel.grid     = element_blank(),
        legend.text    = element_text(face = "italic", size = 8),
        plot.margin    = margin(5, 5, 0, 5))

actual_order_raw <- ggplot_build(p_mv_raw)$layout$panel_params[[1]]$x$limits

annot_raw <- data.frame(sample_data(ps_raw_rel), stringsAsFactors = FALSE) %>%
  rownames_to_column("SampleID") %>%
  filter(SampleID %in% actual_order_raw) %>%
  mutate(SampleID = factor(SampleID, levels = actual_order_raw))

region_cols <- c("Neotropics"             = "grey80",
                 "AfroTropics"            = "grey45",
                 "Indo-Malay-Australasia" = "grey15")

p_region_raw <- ggplot(annot_raw, aes(x = SampleID, y = 1, fill = Region)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_manual(values = region_cols, name = "Region") +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = NULL, y = "Region") +
  theme_void() +
  theme(legend.position  = "right",
        legend.key.size  = unit(0.4, "cm"),
        legend.text      = element_text(size = 8),
        legend.title     = element_text(size = 8, face = "bold"),
        axis.title.y     = element_text(size = 7, angle = 0, vjust = 0.5, hjust = 1),
        plot.margin      = margin(0, 5, 2, 5))

left_col_raw <- cowplot::plot_grid(
  p_mv_raw + theme(legend.position = "none"),
  p_region_raw + theme(legend.position = "none"),
  ncol = 1, align = "v", axis = "lr",
  rel_heights = c(8, 0.5)
)

right_col_raw <- cowplot::plot_grid(
  cowplot::get_legend(p_mv_raw),
  cowplot::get_legend(p_region_raw),
  ncol = 1, rel_heights = c(3, 1)
)

cowplot::plot_grid(left_col_raw, right_col_raw, ncol = 2, rel_widths = c(10, 1.5))

ggsave("figures/raw/composition_barplot_unrarefied.pdf", width = 14, height = 5)
```

## 3.3d Iris plot (circular, original style)

```{r iris-plot, fig.width=10, fig.height=10}
phyloseqOBJ %>%
  tax_filter(min_prevalence = 2.5 / 100, verbose = FALSE) %>%
  tax_transform(rank = "Genus", trans = "clr", zero_replace = "halfmin") %>%
  ord_calc(method = "PCA") %>%
  ord_plot_iris(
    tax_level         = "Genus",
    n_taxa            = 10,
    other             = "Other",
    anno_colour       = "Region",
    anno_colour_style = list(alpha = 0.6, size = 3, show.legend = TRUE)
  )

ggsave("figures/raw/iris_plot.pdf", width = 10, height = 10)
```

## 3.3e Composition barplot horizontal (original style)

```{r composition-barplot-horizontal, fig.width=14, fig.height=10}
phyloseqOBJ %>%
  comp_barplot(tax_level = "Genus", facet_by = "Bee_Genus", n_taxa = 13) +
  coord_flip()

ggsave("figures/raw/composition_barplot_horizontal.pdf", width = 14, height = 10)
```

## 3.4 Top taxa by abundance and prevalence

```{r top-taxa}
ps.genus  <- phyloseqOBJ %>% tax_agg(rank = "Genus")
genus_mat <- as.data.frame(t(ps.genus@otu_table))

# Top genera by total abundance
top_abundance <- data.frame(
  Genus          = colnames(genus_mat),
  TotalAbundance = colSums(genus_mat)
) %>%
  arrange(desc(TotalAbundance)) %>%
  mutate(RelAbundance_pct = TotalAbundance / sum(TotalAbundance) * 100)

# Top genera by prevalence
top_prevalence <- data.frame(
  Genus             = colnames(genus_mat),
  N_Samples_Present = colSums(genus_mat > 0)
) %>%
  arrange(desc(N_Samples_Present)) %>%
  mutate(Prevalence_pct = N_Samples_Present / nrow(genus_mat) * 100)

# Combined table
top_combined <- top_abundance %>%
  left_join(top_prevalence, by = "Genus") %>%
  arrange(desc(TotalAbundance))

cat("Top 20 genera by abundance and prevalence:\n")
print(head(top_combined, 20))
```

## 3.5 Rank-abundance curve

```{r rank-abundance-curve, fig.width=11, fig.height=5}
focal_taxa <- c("Snodgrassella", "Gilliamella", "Lactobacillus",
                "Commensalibacter", "Bifidobacterium",
                "Acetobacteraceae Family", "Lactobacillaceae Family")

# Mean proportion per genus averaged across samples (0–1 scale)
mean_prop <- colMeans(sweep(genus_mat, 1, rowSums(genus_mat), "/"))

rank_df <- data.frame(
  Genus     = names(mean_prop),
  MeanProp  = mean_prop
) %>%
  arrange(desc(MeanProp)) %>%
  slice_head(n = 15) %>%
  mutate(
    Rank  = row_number(),
    Focal = Genus %in% focal_taxa,
    Genus = factor(Genus, levels = Genus)
  )

ggplot(rank_df, aes(x = Genus, y = MeanProp)) +
  geom_point(color = "#7b2d8b", size = 3, alpha = 0.9) +
  labs(x = "Bacterial genera ranked by abundance",
       y = "Mean proportion of sequences\nacross samples") +
  theme_bw(base_size = 11) +
  theme(axis.text.x    = element_text(angle = 45, hjust = 1, face = "italic"),
        legend.position = "top",
        panel.grid.minor = element_blank())

ggsave("figures/raw/rank_abundance_curve.pdf", width = 11, height = 5)
```

---

# 4. PERMANOVA and Mantel Test

## 4.1 PERMANOVA

Tests whether microbial community composition differs significantly across geographic
regions using Bray-Curtis distances (999 permutations).
Note: the `Region` variable (AfroTropics, Indo-Malay-Australasia, Neotropics) corresponds
to "continent" as reported in the manuscript.

```{r permanova}
dists_bray <- tax_transform(phyloseqOBJ, trans = "identity", rank = "Family") %>%
  dist_calc("bray")

bray_perm <- dists_bray %>%
  dist_permanova(
    seed        = 1234,
    n_processes = 2,
    n_perms     = 999,
    variables   = "Region"
  )

perm_get(bray_perm) %>% as.data.frame()
```

---

# 5. Phylogenetic Comparative Methods

This section tests whether the relative abundance of four focal gut symbionts
(*Snodgrassella*, *Gilliamella*, *Lactobacillus*, *Commensalibacter*) is associated
with host colony size, while accounting for shared evolutionary history via
phylogenetic generalized least squares (PGLS).

## 5.1 Prepare symbiont prevalence data

```{r prepare-symbiont-data}
# Aggregate rarefied phyloseq to genus level and convert to relative abundances
Prevalence_Genus <- as.data.frame(make_relative(t(ps.genus@otu_table)))

# Extract four focal symbionts (all identified at genus level)
Prev_Snodgrassella    <- Prevalence_Genus %>% select(Snodgrassella)    %>% rownames_to_column("SampleID")
Prev_Gilliamella      <- Prevalence_Genus %>% select(Gilliamella)      %>% rownames_to_column("SampleID")
Prev_Lactobacillus    <- Prevalence_Genus %>% select(Lactobacillus)    %>% rownames_to_column("SampleID")
Prev_Commensalibacter <- Prevalence_Genus %>% select(Commensalibacter) %>% rownames_to_column("SampleID")

Prev_Symbionts <- Prev_Snodgrassella %>%
  left_join(Prev_Gilliamella,      by = "SampleID") %>%
  left_join(Prev_Lactobacillus,    by = "SampleID") %>%
  left_join(Prev_Commensalibacter, by = "SampleID")

cat("Relative abundance table — first rows:\n")
head(Prev_Symbionts)
```

## 5.2 Load and prepare phylogenetic tree

The phylogeny is from Rasmussen & Cameron (2010). `chronos()` forces it to be
ultrametric, which is required for ancestral state reconstruction.

```{r load-tree}
meli.tree           <- read.nexus(paste0(path_raw, "meliponini5genes.nex"))
meli.tree.one       <- meli.tree[[1]]
meli.tree.one.ultra <- chronos(phy = meli.tree.one, lambda = 1)

plot(meli.tree.one.ultra, cex = 0.4,
     main = expression(paste("Ultrametric tree, ", lambda, " = 1")))
```

```{r alternative-lambda, eval=FALSE}
# --- ALTERNATIVE APPROACH: comparing lambda values (0 to 1) ---
# All values yielded consistent topologies; lambda = 1 was retained.
# bee.tree.ult0  <- chronos(phy = meli.tree.one, lambda = 0)
# bee.tree.ult02 <- chronos(phy = meli.tree.one, lambda = 0.2)
# bee.tree.ult04 <- chronos(phy = meli.tree.one, lambda = 0.4)
# bee.tree.ult06 <- chronos(phy = meli.tree.one, lambda = 0.6)
# bee.tree.ult08 <- chronos(phy = meli.tree.one, lambda = 0.8)
```

## 5.3 Match tree tips to metadata and prune tree

```{r match-and-prune}
# Extract voucher numbers from tip labels
df_tree_voucher <- data.frame(VoucherNumber = meli.tree.one.ultra$tip.label) %>%
  mutate(CollectionID = str_extract(VoucherNumber, "\\d+"))

# Merge symbiont data with metadata and voucher mapping
metadata_symbionts <- merge(
  Prev_Symbionts %>%
    left_join(metadata %>% rename(SampleID = sample), by = "SampleID"),
  df_tree_voucher,
  by.x = "CollectionID", by.y = "CollectionID"
) %>%
  mutate(CollectionID = as.character(CollectionID))

saveRDS(metadata_symbionts, file = paste0(path_proc, "metadata_symbionts.rds"))

cat("Unique bee species:", n_distinct(metadata_symbionts$Species), "\n")

# Prune tree to matched samples
tips_to_keep <- unique(metadata_symbionts$VoucherNumber)
pruned_tree  <- drop.tip(
  meli.tree.one.ultra,
  meli.tree.one.ultra$tip.label[!(meli.tree.one.ultra$tip.label %in% tips_to_keep)]
)

class(pruned_tree) <- "phylo"

cat("Tips in pruned tree:", length(pruned_tree$tip.label), "\n")
plot(pruned_tree, cex = 0.5)

saveRDS(pruned_tree, file = paste0(path_proc, "pruned_tree.rds"))
```

## 5.4 Average symbiont prevalence per voucher and binarize

Symbiont relative abundance is averaged per voucher (species), then converted to
binary presence/absence using a threshold of 0.1%, following the methods described
in the paper.

```{r binarize-symbionts}
averages_Snodgrassella    <- aggregate(Snodgrassella    ~ VoucherNumber, data = metadata_symbionts, FUN = mean)
averages_Gilliamella      <- aggregate(Gilliamella      ~ VoucherNumber, data = metadata_symbionts, FUN = mean)
averages_Lactobacillus    <- aggregate(Lactobacillus    ~ VoucherNumber, data = metadata_symbionts, FUN = mean)
averages_Commensalibacter <- aggregate(Commensalibacter ~ VoucherNumber, data = metadata_symbionts, FUN = mean)

# Threshold of 0.1% (0.001) to minimize false positives from cross-contamination
averages_Snodgrassella$Discrete_Snodgrassella       <- as.numeric(averages_Snodgrassella$Snodgrassella       > 0.001)
averages_Gilliamella$Discrete_Gilliamella           <- as.numeric(averages_Gilliamella$Gilliamella           > 0.001)
averages_Lactobacillus$Discrete_Lactobacillus       <- as.numeric(averages_Lactobacillus$Lactobacillus       > 0.001)
averages_Commensalibacter$Discrete_Commensalibacter <- as.numeric(averages_Commensalibacter$Commensalibacter > 0.001)

discrete_Sng   <- averages_Snodgrassella[,    c("VoucherNumber", "Discrete_Snodgrassella")]
discrete_Gill  <- averages_Gilliamella[,      c("VoucherNumber", "Discrete_Gilliamella")]
discrete_Lacto <- averages_Lactobacillus[,    c("VoucherNumber", "Discrete_Lactobacillus")]
discrete_Comm  <- averages_Commensalibacter[, c("VoucherNumber", "Discrete_Commensalibacter")]

saveRDS(discrete_Sng,   file = paste0(path_proc, "discrete_Sng.rds"))
saveRDS(discrete_Gill,  file = paste0(path_proc, "discrete_Gill.rds"))
saveRDS(discrete_Lacto, file = paste0(path_proc, "discrete_Lacto.rds"))
saveRDS(discrete_Comm,  file = paste0(path_proc, "discrete_Comm.rds"))
```

## 5.5 Prepare PGLS data frames

```{r prepare-pgls}
df_subset <- metadata_symbionts[, c("VoucherNumber", "ColonySize_U")] %>%
  distinct()

averages_merged <- merge(df_subset, averages_Snodgrassella, by = "VoucherNumber") %>%
  merge(averages_Gilliamella,      by = "VoucherNumber") %>%
  merge(averages_Lactobacillus,    by = "VoucherNumber") %>%
  merge(averages_Commensalibacter, by = "VoucherNumber")

# Remove duplicates (PGLS requires one row per tree tip)
unique_all_avg <- unique(averages_merged)

cat("Rows in merged table:", nrow(averages_merged),
    "| Unique rows:", nrow(unique_all_avg), "\n")

saveRDS(unique_all_avg, file = paste0(path_proc, "unique_all_avg.rds"))
```

## 5.6 PGLS: symbiont relative abundance vs. colony size

```{r pgls-models}
# Base alignment: match all 29 species to tree, add log colony size
unique_all_avg <- unique_all_avg %>%
  mutate(ColonySize_log = log(ColonySize_U)) %>%
  filter(!is.na(ColonySize_log), !is.na(VoucherNumber))

common      <- intersect(pruned_tree$tip.label, unique_all_avg$VoucherNumber)
pruned_pgls <- drop.tip(pruned_tree, setdiff(pruned_tree$tip.label, common))
unique_all_avg <- unique_all_avg[match(common, unique_all_avg$VoucherNumber), ]

# PGLS uses all species with any detectable symbiont (> 0); the 0.1% threshold is
# reserved for binary presence/absence classification (Pagel's test, ancestral reconstruction).
# Species with zero abundance are shown at 1e-04 floor in the plot for visibility on log scale.

# Snodgrassella vs. Colony Size
snod_df <- unique_all_avg %>%
  filter(Snodgrassella > 0) %>%
  mutate(Snodgrassella_log = log(Snodgrassella))
common_snod  <- intersect(pruned_pgls$tip.label, snod_df$VoucherNumber)
pruned_snod  <- drop.tip(pruned_pgls, setdiff(pruned_pgls$tip.label, common_snod))
snod_df      <- snod_df[match(common_snod, snod_df$VoucherNumber), ]
spp_snod     <- snod_df$VoucherNumber
corBM_snod   <- corBrownian(phy = pruned_snod, form = ~spp_snod)
pgls_Sn_CS   <- gls(Snodgrassella_log ~ ColonySize_log,
                     data = snod_df, correlation = corBM_snod)
summary(pgls_Sn_CS)

# Gilliamella vs. Colony Size
gill_df <- unique_all_avg %>%
  filter(Gilliamella > 0) %>%
  mutate(Gilliamella_log = log(Gilliamella))
common_gill  <- intersect(pruned_pgls$tip.label, gill_df$VoucherNumber)
pruned_gill  <- drop.tip(pruned_pgls, setdiff(pruned_pgls$tip.label, common_gill))
gill_df      <- gill_df[match(common_gill, gill_df$VoucherNumber), ]
spp_gill     <- gill_df$VoucherNumber
corBM_gill   <- corBrownian(phy = pruned_gill, form = ~spp_gill)
pgls_Gill_CS <- gls(Gilliamella_log ~ ColonySize_log,
                     data = gill_df, correlation = corBM_gill)
summary(pgls_Gill_CS)

# Lactobacillus vs. Colony Size
lacto_df <- unique_all_avg %>%
  filter(Lactobacillus > 0) %>%
  mutate(Lactobacillus_log = log(Lactobacillus))
common_lacto  <- intersect(pruned_pgls$tip.label, lacto_df$VoucherNumber)
pruned_lacto  <- drop.tip(pruned_pgls, setdiff(pruned_pgls$tip.label, common_lacto))
lacto_df      <- lacto_df[match(common_lacto, lacto_df$VoucherNumber), ]
spp_lacto     <- lacto_df$VoucherNumber
corBM_lacto   <- corBrownian(phy = pruned_lacto, form = ~spp_lacto)
pgls_Lacto_CS <- gls(Lactobacillus_log ~ ColonySize_log,
                      data = lacto_df, correlation = corBM_lacto)
summary(pgls_Lacto_CS)

# Commensalibacter vs. Colony Size
comm_df <- unique_all_avg %>%
  filter(Commensalibacter > 0) %>%
  mutate(Commensalibacter_log = log(Commensalibacter))
common_comm  <- intersect(pruned_pgls$tip.label, comm_df$VoucherNumber)
pruned_comm  <- drop.tip(pruned_pgls, setdiff(pruned_pgls$tip.label, common_comm))
comm_df      <- comm_df[match(common_comm, comm_df$VoucherNumber), ]
spp_comm     <- comm_df$VoucherNumber
corBM_comm   <- corBrownian(phy = pruned_comm, form = ~spp_comm)
pgls_Comm_CS <- gls(Commensalibacter_log ~ ColonySize_log,
                     data = comm_df, correlation = corBM_comm)
summary(pgls_Comm_CS)
```

## 5.7 Visualize PGLS results

```{r pgls-plots, fig.width=10, fig.height=5}
# All 29 species shown; zeros mapped to 1e-04 floor for visibility on log scale.
# Smoothing line and R\u00b2 include all species (with floor values), matching the full dataset.
unique_all_avg <- unique_all_avg %>%
  mutate(
    Snod_floor  = pmax(Snodgrassella,    1e-04),
    Gill_floor  = pmax(Gilliamella,      1e-04),
    Lacto_floor = pmax(Lactobacillus,    1e-04),
    Comm_floor  = pmax(Commensalibacter, 1e-04)
  )

make_pgls_plot_rel <- function(df, floor_col, mdl, fill_col, label) {
  p_val <- summary(mdl)$tTable[2, 4]
  r2    <- summary(lm(as.formula(paste("log(", floor_col, ") ~ log(ColonySize_U)")),
                      data = df))$r.squared
  ggplot(df, aes(x = ColonySize_U, y = .data[[floor_col]])) +
    geom_point(shape = 21, fill = fill_col, size = 3, alpha = 0.8) +
    geom_smooth(method = "lm", se = FALSE, color = "grey", linetype = "dashed") +
    scale_x_log10() + scale_y_log10() +
    annotation_logticks() +
    annotate("text", x = Inf, y = Inf,
             label = paste("PGLS p =", signif(p_val, 3), "\nR\u00b2 =", round(r2, 2)),
             hjust = 1.1, vjust = 1.5, size = 3.5, fontface = "italic") +
    labs(x = "Colony Size",
         y = bquote(italic(.(label)) ~ "proportion")) +
    theme_bw() +
    theme(panel.grid.minor = element_blank(), panel.border = element_blank(),
          axis.line = element_line(colour = "black"))
}

SnodPlot  <- make_pgls_plot_rel(unique_all_avg, "Snod_floor",  pgls_Sn_CS,   "#e54b4b", "Snodgrassella")
GillPlot  <- make_pgls_plot_rel(unique_all_avg, "Gill_floor",  pgls_Gill_CS, "#2b9dce", "Gilliamella")
LactoPlot <- make_pgls_plot_rel(unique_all_avg, "Lacto_floor", pgls_Lacto_CS,"#6ab187", "Lactobacillus")
CommPlot  <- make_pgls_plot_rel(unique_all_avg, "Comm_floor",  pgls_Comm_CS, "#f0a500", "Commensalibacter")

grid.arrange(SnodPlot, GillPlot, LactoPlot, CommPlot, nrow = 1)
ggsave("figures/raw/PGLS_relAbundance_colonySize.pdf", width = 18, height = 5)
```

---

## 5.8 Symbiont relative abundance by bee genus

```{r symbiont-boxplots-genus, fig.width=14, fig.height=8}
symbiont_colors <- c("Snodgrassella"    = "#e54b4b",
                     "Gilliamella"      = "#2b9dce",
                     "Lactobacillus"    = "#6ab187",
                     "Commensalibacter" = "#f0a500")

# Pivot to long format
symb_long <- metadata_symbionts %>%
  select(SampleID, Genus, Snodgrassella, Gilliamella,
         Lactobacillus, Commensalibacter) %>%
  pivot_longer(cols      = c(Snodgrassella, Gilliamella,
                              Lactobacillus, Commensalibacter),
               names_to  = "Symbiont",
               values_to = "RelAbun") %>%
  mutate(Symbiont = factor(Symbiont, levels = names(symbiont_colors)))

# Order bee genera by median Snodgrassella relative abundance
genus_order_symb <- symb_long %>%
  filter(Symbiont == "Snodgrassella") %>%
  group_by(Genus) %>%
  summarise(med = median(RelAbun, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(med)) %>%
  pull(Genus)

symb_long$Genus <- factor(symb_long$Genus, levels = genus_order_symb)

ggplot(symb_long, aes(x = Genus, y = RelAbun + 1e-04,
                       fill = Symbiont, color = Symbiont)) +
  geom_boxplot(alpha = 0.35, outlier.shape = NA, width = 0.6) +
  geom_jitter(width = 0.15, size = 0.7, alpha = 0.5) +
  scale_y_log10(labels = scales::percent_format(accuracy = 0.01)) +
  scale_fill_manual(values  = symbiont_colors) +
  scale_color_manual(values = symbiont_colors) +
  facet_wrap(~ Symbiont, nrow = 2, scales = "free_y") +
  labs(x = NULL, y = "Relative Abundance (log scale)") +
  theme_bw(base_size = 11) +
  theme(axis.text.x      = element_text(angle = 45, hjust = 1, face = "italic"),
        panel.grid.minor = element_blank(),
        strip.text       = element_text(face = "italic"),
        legend.position  = "none")

ggsave("figures/raw/symbiont_relAbun_byGenus_boxplots.pdf", width = 14, height = 8)
```

---

# 6. Pagel's Binary Trait Correlation Tests

## 6.1 Prepare categorical data

```{r pagel-prep}
# Merge binary presence/absence for all four symbionts into a single table.
# make.unique() ensures row names are unique in case any VoucherNumber appears
# more than once after merging (required by plotTree.datamatrix and fitPagel).
discrete_all <- merge(discrete_Sng,   discrete_Gill,  by = "VoucherNumber") %>%
  merge(discrete_Lacto, by = "VoucherNumber") %>%
  merge(discrete_Comm,  by = "VoucherNumber") %>%
  mutate(VoucherNumber = make.unique(as.character(VoucherNumber))) %>%
  column_to_rownames(var = "VoucherNumber")


averages_merged_filtered_unique <- discrete_all %>%
  mutate(
    Discrete_Snodgrassella    = factor(Discrete_Snodgrassella,    levels = c(0,1), labels = c("S0",  "S1")),
    Discrete_Gilliamella      = factor(Discrete_Gilliamella,      levels = c(0,1), labels = c("G0",  "G1")),
    Discrete_Lactobacillus    = factor(Discrete_Lactobacillus,    levels = c(0,1), labels = c("L0",  "L1")),
    Discrete_Commensalibacter = factor(Discrete_Commensalibacter, levels = c(0,1), labels = c("C0",  "C1"))
  )

# Plot phylogeny with binary trait matrix for all four symbionts
Binary_Symbionts <- plotTree.datamatrix(
  pruned_tree,
  averages_merged_filtered_unique,
  fsize = 0.5, yexp = 1, header = FALSE, xexp = 1.45,
  palettes = c("Reds", "Blues", "Greens", "Oranges")
)
```

## 6.2 Snodgrassella vs. Gilliamella

```{r pagel-snod-gill}
Snodgrassella_Dis_mode <- setNames(averages_merged_filtered_unique[, 1],
                                    rownames(averages_merged_filtered_unique))
Gilliamella_Dis_mode      <- setNames(averages_merged_filtered_unique[, 2],
                                    rownames(averages_merged_filtered_unique))

Symbionts.Dis.fit <- fitPagel(pruned_tree, Snodgrassella_Dis_mode, Gilliamella_Dis_mode)
print(Symbionts.Dis.fit)

plot(Symbionts.Dis.fit, signif = 4, cex.main = 1, cex.sub = 0.8,
     cex.traits = 0.7, cex.rates = 0.7, lwd = 1)
```

## 6.3 Snodgrassella vs. colony size

```{r pagel-snod-colonysize}
median_cs <- median(unique_all_avg$ColonySize_U, na.rm = TRUE)

discrete_snod_colsize <- unique_all_avg %>%
  select(VoucherNumber, Discrete_Snodgrassella, ColonySize_U) %>%
  mutate(Discrete_ColonySize = ifelse(ColonySize_U > median_cs, "Large", "Small")) %>%
  select(VoucherNumber, Discrete_Snodgrassella, Discrete_ColonySize) %>%
  as.data.frame() %>%
  remove_rownames() %>%
  column_to_rownames(var = "VoucherNumber")

Snodgrassella_Dis_mode <- setNames(discrete_snod_colsize$Discrete_Snodgrassella,
                                    rownames(discrete_snod_colsize))
ColonySize_Dis_mode    <- setNames(discrete_snod_colsize$Discrete_ColonySize,
                                    rownames(discrete_snod_colsize))

Snod.Cs.Di.fit <- fitPagel(pruned_tree, Snodgrassella_Dis_mode, ColonySize_Dis_mode)
print(Snod.Cs.Di.fit)
plot(Snod.Cs.Di.fit, signif = 4, cex.main = 1, cex.sub = 0.8,
     cex.traits = 0.7, cex.rates = 0.7, lwd = 1)
```

## 6.4 Gilliamella vs. colony size

```{r pagel-gill-colonysize}
discrete_gill_colsize <- unique_all_avg %>%
  select(VoucherNumber, Discrete_Gilliamella, ColonySize_U) %>%
  mutate(Discrete_ColonySize = ifelse(ColonySize_U > median_cs, "Large", "Small")) %>%
  select(VoucherNumber, Discrete_Gilliamella, Discrete_ColonySize) %>%
  as.data.frame() %>%
  remove_rownames() %>%
  column_to_rownames(var = "VoucherNumber")

Gilliamella_Dis_mode   <- setNames(discrete_gill_colsize$Discrete_Gilliamella,
                                 rownames(discrete_gill_colsize))
ColonySize_Dis_mode <- setNames(discrete_gill_colsize$Discrete_ColonySize,
                                 rownames(discrete_gill_colsize))

Gilli.Cs.Di.fit <- fitPagel(pruned_tree, Gilliamella_Dis_mode, ColonySize_Dis_mode)
print(Gilli.Cs.Di.fit)
plot(Gilli.Cs.Di.fit, signif = 4, cex.main = 1, cex.sub = 0.8,
     cex.traits = 0.7, cex.rates = 0.7, lwd = 1)
```

## 6.5 Lactobacillus vs. colony size

```{r pagel-lacto-colonysize}
discrete_lacto_colsize <- unique_all_avg %>%
  select(VoucherNumber, Discrete_Lactobacillus, ColonySize_U) %>%
  mutate(Discrete_ColonySize = ifelse(ColonySize_U > median_cs, "Large", "Small")) %>%
  select(VoucherNumber, Discrete_Lactobacillus, Discrete_ColonySize) %>%
  as.data.frame() %>%
  remove_rownames() %>%
  column_to_rownames(var = "VoucherNumber")

Lactobacillus_Dis_mode <- setNames(discrete_lacto_colsize$Discrete_Lactobacillus,
                                    rownames(discrete_lacto_colsize))
ColonySize_Dis_mode    <- setNames(discrete_lacto_colsize$Discrete_ColonySize,
                                    rownames(discrete_lacto_colsize))

if (length(unique(Lactobacillus_Dis_mode)) < 2) {
  cat("Lactobacillus is monomorphic across tree species (all present or all absent).",
      "\nPagel's test is not applicable.\n")
} else {
  Lacto.Cs.Di.fit <- fitPagel(pruned_tree, Lactobacillus_Dis_mode, ColonySize_Dis_mode)
  print(Lacto.Cs.Di.fit)
  plot(Lacto.Cs.Di.fit, signif = 4, cex.main = 1, cex.sub = 0.8,
       cex.traits = 0.7, cex.rates = 0.7, lwd = 1)
}
```

## 6.6 Commensalibacter vs. colony size

```{r pagel-comm-colonysize}
discrete_comm_colsize <- unique_all_avg %>%
  select(VoucherNumber, Discrete_Commensalibacter, ColonySize_U) %>%
  mutate(Discrete_ColonySize = ifelse(ColonySize_U > median_cs, "Large", "Small")) %>%
  select(VoucherNumber, Discrete_Commensalibacter, Discrete_ColonySize) %>%
  as.data.frame() %>%
  remove_rownames() %>%
  column_to_rownames(var = "VoucherNumber")

Commensalibacter_Dis_mode <- setNames(discrete_comm_colsize$Discrete_Commensalibacter,
                                       rownames(discrete_comm_colsize))
ColonySize_Dis_mode       <- setNames(discrete_comm_colsize$Discrete_ColonySize,
                                       rownames(discrete_comm_colsize))

if (length(unique(Commensalibacter_Dis_mode)) < 2) {
  cat("Commensalibacter is monomorphic across tree species (all present or all absent).",
      "\nPagel's test is not applicable.\n")
} else {
  Comm.Cs.Di.fit <- fitPagel(pruned_tree, Commensalibacter_Dis_mode, ColonySize_Dis_mode)
  print(Comm.Cs.Di.fit)
  plot(Comm.Cs.Di.fit, signif = 4, cex.main = 1, cex.sub = 0.8,
       cex.traits = 0.7, cex.rates = 0.7, lwd = 1)
}
```

---

# 7. Phylogenetic ANOVA

```{r phylo-anova}
discrete_all$Discrete_Snodgrassella    <- as.factor(discrete_all$Discrete_Snodgrassella)
discrete_all$Discrete_Gilliamella      <- as.factor(discrete_all$Discrete_Gilliamella)
discrete_all$Discrete_Lactobacillus    <- as.factor(discrete_all$Discrete_Lactobacillus)
discrete_all$Discrete_Commensalibacter <- as.factor(discrete_all$Discrete_Commensalibacter)

anova_snod <- phylANOVA(pruned_tree,
                         discrete_all$Discrete_Snodgrassella,
                         averages_merged$ColonySize_U,
                         nsim = 1000)

anova_gill  <- phylANOVA(pruned_tree,
                         discrete_all$Discrete_Gilliamella,
                         averages_merged$ColonySize_U,
                         nsim = 1000)

print(anova_snod)
print(anova_gill)

if (length(levels(discrete_all$Discrete_Lactobacillus)) >= 2) {
  anova_lacto <- phylANOVA(pruned_tree,
                            discrete_all$Discrete_Lactobacillus,
                            averages_merged$ColonySize_U,
                            nsim = 1000)
  print(anova_lacto)
} else {
  cat("Lactobacillus is monomorphic — phylANOVA not applicable.\n")
  anova_lacto <- NULL
}

if (length(levels(discrete_all$Discrete_Commensalibacter)) >= 2) {
  anova_comm <- phylANOVA(pruned_tree,
                           discrete_all$Discrete_Commensalibacter,
                           averages_merged$ColonySize_U,
                           nsim = 1000)
  print(anova_comm)
} else {
  cat("Commensalibacter is monomorphic — phylANOVA not applicable.\n")
  anova_comm <- NULL
}
```

```{r phylo-anova-plots, fig.width=10, fig.height=5}
# Add colony size to binary symbiont table for plotting
averages_merged_filtered_unique$VoucherNumber <- rownames(averages_merged_filtered_unique)
averages_merged_bs_unique <- averages_merged[!duplicated(averages_merged$VoucherNumber), ]

averages_merged_filtered_unique_CS <- merge(
  averages_merged_filtered_unique,
  averages_merged_bs_unique[, c("VoucherNumber", "ColonySize_U")],
  by = "VoucherNumber"
) %>% column_to_rownames(var = "VoucherNumber")

Discrete_Snod_plot <- averages_merged_filtered_unique_CS %>%
  ggplot(aes(x = ColonySize_U,
             y = factor(Discrete_Snodgrassella, labels = c("Absent", "Present")),
             fill = factor(Discrete_Snodgrassella))) +
  geom_boxplot(width = 0.5, alpha = 0.5, outlier.shape = NA) +
  geom_jitter(aes(color = factor(Discrete_Snodgrassella)),
              width = 0.1, height = 0.1, alpha = 0.6, size = 1.5) +
  scale_x_log10() +
  scale_fill_manual(values  = c("white", "#e54b4b")) +
  scale_color_manual(values = c("grey",  "#e54b4b")) +
  labs(x = "Colony Size (log scale)", y = "Snodgrassella Presence") +
  theme_minimal() +
  theme(legend.position = "none") +
  coord_flip()

Discrete_Gill_plot <- averages_merged_filtered_unique_CS %>%
  ggplot(aes(x = ColonySize_U,
             y = factor(Discrete_Gilliamella, labels = c("Absent", "Present")),
             fill = factor(Discrete_Gilliamella))) +
  geom_boxplot(width = 0.5, alpha = 0.5, outlier.shape = NA) +
  geom_jitter(aes(color = factor(Discrete_Gilliamella)),
              width = 0.1, height = 0.1, alpha = 0.6, size = 1.5) +
  scale_x_log10() +
  scale_fill_manual(values  = c("white", "#2b9dce")) +
  scale_color_manual(values = c("grey",  "#2b9dce")) +
  labs(x = "Colony Size (log scale)", y = "Gilliamella Presence") +
  theme_minimal() +
  theme(legend.position = "none") +
  coord_flip()

plot_list <- list(SnodPlot, Discrete_Snod_plot, GillPlot, Discrete_Gill_plot)

if (!is.null(anova_lacto)) {
  Discrete_Lacto_plot <- averages_merged_filtered_unique_CS %>%
    ggplot(aes(x = ColonySize_U,
               y = factor(Discrete_Lactobacillus, labels = c("Absent", "Present")),
               fill = factor(Discrete_Lactobacillus))) +
    geom_boxplot(width = 0.5, alpha = 0.5, outlier.shape = NA) +
    geom_jitter(aes(color = factor(Discrete_Lactobacillus)),
                width = 0.1, height = 0.1, alpha = 0.6, size = 1.5) +
    scale_x_log10() +
    scale_fill_manual(values  = c("white", "#6ab187")) +
    scale_color_manual(values = c("grey",  "#6ab187")) +
    labs(x = "Colony Size (log scale)", y = "Lactobacillus Presence") +
    theme_minimal() +
    theme(legend.position = "none") +
    coord_flip()
  plot_list <- c(plot_list, list(LactoPlot, Discrete_Lacto_plot))
}

if (!is.null(anova_comm)) {
  Discrete_Comm_plot <- averages_merged_filtered_unique_CS %>%
    ggplot(aes(x = ColonySize_U,
               y = factor(Discrete_Commensalibacter, labels = c("Absent", "Present")),
               fill = factor(Discrete_Commensalibacter))) +
    geom_boxplot(width = 0.5, alpha = 0.5, outlier.shape = NA) +
    geom_jitter(aes(color = factor(Discrete_Commensalibacter)),
                width = 0.1, height = 0.1, alpha = 0.6, size = 1.5) +
    scale_x_log10() +
    scale_fill_manual(values  = c("white", "#f0a500")) +
    scale_color_manual(values = c("grey",  "#f0a500")) +
    labs(x = "Colony Size (log scale)", y = "Commensalibacter Presence") +
    theme_minimal() +
    theme(legend.position = "none") +
    coord_flip()
  plot_list <- c(plot_list, list(CommPlot, Discrete_Comm_plot))
}

do.call(grid.arrange, c(plot_list, list(nrow = length(plot_list) / 2)))
ggsave("figures/raw/phyloANOVA_colonySize.pdf", width = 10,
       height = 4 * length(plot_list) / 2)
```

---

# 8. Ancestral State Reconstruction

## 8.1 Model selection

```{r model-selection}
snod.mode  <- setNames(as.factor(discrete_Sng[, 2]),   discrete_Sng$VoucherNumber)
gill.mode   <- setNames(as.factor(discrete_Gill[, 2]),  discrete_Gill$VoucherNumber)
lacto.mode <- setNames(as.factor(discrete_Lacto[, 2]), discrete_Lacto$VoucherNumber)
comm.mode  <- setNames(as.factor(discrete_Comm[, 2]),  discrete_Comm$VoucherNumber)

cols_snod  <- setNames(c("white", "salmon"),   levels(snod.mode))
cols_gill   <- setNames(c("white", "#2b9dce"),  levels(gill.mode))
cols_lacto <- setNames(c("white", "#6ab187"),  levels(lacto.mode))
cols_comm  <- setNames(c("white", "#f0a500"),  levels(comm.mode))

fit_models <- function(tree, trait) {
  fitER  <- fitMk(tree, trait, model = "ER")
  fitARD <- fitMk(tree, trait, model = "ARD")
  fit01  <- fitMk(tree, trait, model = matrix(c(0,1,0,0), 2, 2, byrow = TRUE))
  fit10  <- fitMk(tree, trait, model = matrix(c(0,0,1,0), 2, 2, byrow = TRUE))
  aic    <- c(AIC(fitER), AIC(fitARD), AIC(fit01), AIC(fit10))
  data.frame(
    model     = c("ER", "ARD", "P->A", "A->P"),
    logL      = c(logLik(fitER), logLik(fitARD), logLik(fit01), logLik(fit10)),
    AIC       = aic,
    delta.AIC = aic - min(aic)
  )
}

cat("Model selection — Snodgrassella:\n")
print(fit_models(pruned_tree, snod.mode))

cat("\nModel selection — Gilliamella:\n")
print(fit_models(pruned_tree, gill.mode))

if (length(levels(lacto.mode)) >= 2) {
  cat("\nModel selection — Lactobacillus:\n")
  print(fit_models(pruned_tree, lacto.mode))
} else {
  cat("\nLactobacillus is monomorphic — model selection not applicable.\n")
}

if (length(levels(comm.mode)) >= 2) {
  cat("\nModel selection — Commensalibacter:\n")
  print(fit_models(pruned_tree, comm.mode))
} else {
  cat("\nCommensalibacter is monomorphic — model selection not applicable.\n")
}
```

## 8.2 Marginal ancestral reconstruction (corHMM, ER model)

```{r ancestral-reconstruction}
fit.marginal       <- corHMM(pruned_tree, discrete_Sng,   node.states = "marginal",
                              rate.cat = 1, rate.mat = matrix(c(NA,1,1,NA), 2, 2))
fit.marginal_gill   <- corHMM(pruned_tree, discrete_Gill,  node.states = "marginal",
                              rate.cat = 1, rate.mat = matrix(c(NA,1,1,NA), 2, 2))
# Snodgrassella
plotTree.datamatrix(pruned_tree, as.data.frame(snod.mode),
                    colors = list(cols_snod), header = FALSE, fsize = 0.45)
legend("topright", legend = levels(snod.mode), pch = 22, pt.cex = 1.5,
       pt.bg = cols_snod, bty = "n", cex = 0.8)
nodelabels(pie = fit.marginal$states, piecol = cols_snod, cex = 0.5)

# Gilliamella
plotTree.datamatrix(pruned_tree, as.data.frame(gill.mode),
                    colors = list(cols_gill), header = FALSE, fsize = 0.45)
legend("topright", legend = levels(gill.mode), pch = 22, pt.cex = 1.5,
       pt.bg = cols_gill, bty = "n", cex = 0.8)
nodelabels(pie = fit.marginal_gill$states, piecol = cols_gill, cex = 0.5)

# Lactobacillus
if (length(levels(lacto.mode)) >= 2) {
  fit.marginal_lacto <- corHMM(pruned_tree, discrete_Lacto, node.states = "marginal",
                                rate.cat = 1, rate.mat = matrix(c(NA,1,1,NA), 2, 2))
  plotTree.datamatrix(pruned_tree, as.data.frame(lacto.mode),
                      colors = list(cols_lacto), header = FALSE, fsize = 0.45)
  legend("topright", legend = levels(lacto.mode), pch = 22, pt.cex = 1.5,
         pt.bg = cols_lacto, bty = "n", cex = 0.8)
  nodelabels(pie = fit.marginal_lacto$states, piecol = cols_lacto, cex = 0.5)
} else {
  cat("Lactobacillus is monomorphic — ancestral reconstruction not applicable.\n")
  fit.marginal_lacto <- NULL
}

# Commensalibacter
if (length(levels(comm.mode)) >= 2) {
  fit.marginal_comm <- corHMM(pruned_tree, discrete_Comm,  node.states = "marginal",
                               rate.cat = 1, rate.mat = matrix(c(NA,1,1,NA), 2, 2))
  plotTree.datamatrix(pruned_tree, as.data.frame(comm.mode),
                      colors = list(cols_comm), header = FALSE, fsize = 0.45)
  legend("topright", legend = levels(comm.mode), pch = 22, pt.cex = 1.5,
         pt.bg = cols_comm, bty = "n", cex = 0.8)
  nodelabels(pie = fit.marginal_comm$states, piecol = cols_comm, cex = 0.5)
} else {
  cat("Commensalibacter is monomorphic — ancestral reconstruction not applicable.\n")
  fit.marginal_comm <- NULL
}
```

## 8.3 Stochastic character mapping (1,000 simulations)

```{r stochastic-mapping, cache=TRUE}
set.seed(42)

mtrees_snd <- make.simmap(pruned_tree, snod.mode, model = "ER",
                           nsim = 1000, Q = "mcmc", vQ = 0.01,
                           prior = list(use.empirical = TRUE), samplefreq = 10)

mtrees_gill <- make.simmap(pruned_tree, gill.mode, model = "ER",
                           nsim = 1000, Q = "mcmc", vQ = 0.01,
                           prior = list(use.empirical = TRUE), samplefreq = 10)

pd_snd <- summary(mtrees_snd)
pd_gill <- summary(mtrees_gill)

if (length(levels(lacto.mode)) >= 2) {
  mtrees_lacto <- make.simmap(pruned_tree, lacto.mode, model = "ER",
                               nsim = 1000, Q = "mcmc", vQ = 0.01,
                               prior = list(use.empirical = TRUE), samplefreq = 10)
  pd_lacto <- summary(mtrees_lacto)
} else {
  cat("Lactobacillus is monomorphic — stochastic mapping not applicable.\n")
  mtrees_lacto <- NULL; pd_lacto <- NULL
}

if (length(levels(comm.mode)) >= 2) {
  mtrees_comm <- make.simmap(pruned_tree, comm.mode, model = "ER",
                              nsim = 1000, Q = "mcmc", vQ = 0.01,
                              prior = list(use.empirical = TRUE), samplefreq = 10)
  pd_comm <- summary(mtrees_comm)
} else {
  cat("Commensalibacter is monomorphic — stochastic mapping not applicable.\n")
  mtrees_comm <- NULL; pd_comm <- NULL
}
```

```{r stochastic-mapping-plots, fig.width=14, fig.height=16}
par(mfcol = c(2, 2))

plot(pd_snd, colors = cols_snod, fsize = 0.8, ftype = "i", lwd = 2,
     offset = 0.4, cex = c(0.7, 0.5))
legend("bottomleft", legend = levels(snod.mode), pch = 22, pt.cex = 1.5,
       pt.bg = cols_snod, bty = "n", cex = 0.8)

plot(pd_gill, direction = "leftwards", colors = cols_gill, fsize = 0.8,
     ftype = "i", lwd = 2, offset = 0.4, cex = c(0.7, 0.5))
legend("bottomleft", legend = levels(gill.mode), pch = 22, pt.cex = 1.5,
       pt.bg = cols_gill, bty = "n", cex = 0.8)

if (!is.null(pd_lacto)) {
  plot(pd_lacto, colors = cols_lacto, fsize = 0.8, ftype = "i", lwd = 2,
       offset = 0.4, cex = c(0.7, 0.5))
  legend("bottomleft", legend = levels(lacto.mode), pch = 22, pt.cex = 1.5,
         pt.bg = cols_lacto, bty = "n", cex = 0.8)
}

if (!is.null(pd_comm)) {
  plot(pd_comm, direction = "leftwards", colors = cols_comm, fsize = 0.8,
       ftype = "i", lwd = 2, offset = 0.4, cex = c(0.7, 0.5))
  legend("bottomleft", legend = levels(comm.mode), pch = 22, pt.cex = 1.5,
         pt.bg = cols_comm, bty = "n", cex = 0.8)
}
```

```{r root-probabilities}
tips_all  <- pruned_tree$tip.label
mrca_root <- getMRCA(pruned_tree, tips_all)

pp_snd_root <- pd_snd$ace[as.character(mrca_root), ]
pp_gill_root <- pd_gill$ace[as.character(mrca_root), ]

sprintf("Root — Snodgrassella:    absent = %.2f, present = %.2f",
        pp_snd_root[1], pp_snd_root[2])
sprintf("Root — Gilliamella:      absent = %.2f, present = %.2f",
        pp_gill_root[1], pp_gill_root[2])
if (!is.null(pd_lacto)) {
  pp_lacto_root <- pd_lacto$ace[as.character(mrca_root), ]
  sprintf("Root — Lactobacillus:    absent = %.2f, present = %.2f",
          pp_lacto_root[1], pp_lacto_root[2])
} else {
  cat("Root — Lactobacillus: monomorphic, root probability not estimated.\n")
}

if (!is.null(pd_comm)) {
  pp_comm_root <- pd_comm$ace[as.character(mrca_root), ]
  sprintf("Root — Commensalibacter: absent = %.2f, present = %.2f",
          pp_comm_root[1],  pp_comm_root[2])
} else {
  cat("Root — Commensalibacter: monomorphic, root probability not estimated.\n")
}
```

---

# 9. Phylogenetic Signal (Blomberg's K)

```{r phylo-signal}
df_tree_voucher2 <- data.frame(VoucherNumber = meli.tree.one.ultra$tip.label) %>%
  mutate(CollectionID = str_extract(VoucherNumber, "\\d+"))

signal_df <- as.data.frame(t(ps.genus@otu_table)) %>%
  rownames_to_column("SampleID") %>%
  select(SampleID, Snodgrassella, Lactobacillus, Commensalibacter) %>%
  left_join(metadata %>%
              rename(SampleID = sample) %>%
              mutate(CollectionID = as.character(CollectionID)),
            by = "SampleID") %>%
  left_join(df_tree_voucher2, by = "CollectionID") %>%
  filter(!is.na(VoucherNumber)) %>%
  group_by(VoucherNumber) %>%
  summarise(
    Snodgrassella    = mean(Snodgrassella,    na.rm = TRUE),
    Lactobacillus    = mean(Lactobacillus,    na.rm = TRUE),
    Commensalibacter = mean(Commensalibacter, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  as.data.frame()

rownames(signal_df) <- signal_df$VoucherNumber

common_sig  <- intersect(pruned_tree$tip.label, signal_df$VoucherNumber)
tree_sig    <- drop.tip(pruned_tree, setdiff(pruned_tree$tip.label, common_sig))
signal_df   <- signal_df[tree_sig$tip.label, ]
class(tree_sig) <- "phylo"

taxa_to_test <- c("Snodgrassella", "Lactobacillus", "Commensalibacter")

results_signal <- lapply(taxa_to_test, function(taxon) {
  x <- setNames(signal_df[[taxon]], rownames(signal_df))
  k <- phylosig(tree_sig, x, method = "K", test = TRUE, nsim = 1000)
  data.frame(Taxon = taxon, K = round(k$K, 3), P_value = round(k$P, 4))
}) %>% bind_rows()

print(results_signal)
```

---

# 10. Alpha Diversity: Phylogenetic Signal and PGLS

## 10.1 Prepare species-level alpha diversity data

```{r alpha-div-prep}
df_tree_ult1 <- data.frame(VoucherNumber = meli.tree.one.ultra$tip.label) %>%
  mutate(CollectionID = str_extract(VoucherNumber, "\\d+"))

alpha_pgls_df <- alpha.table %>%
  rename(SampleID = Row.names) %>%
  mutate(CollectionID = as.character(CollectionID)) %>%
  left_join(df_tree_ult1, by = "CollectionID") %>%
  filter(!is.na(VoucherNumber)) %>%
  group_by(VoucherNumber) %>%
  summarise(
    Shannon_mean      = mean(Shannon,      na.rm = TRUE),
    ColonySize_U_mean = mean(ColonySize_U, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(Shannon_mean), !is.na(ColonySize_U_mean)) %>%
  as.data.frame()

rownames(alpha_pgls_df) <- alpha_pgls_df$VoucherNumber

common_alpha  <- intersect(pruned_tree$tip.label, alpha_pgls_df$VoucherNumber)
tree_alpha    <- drop.tip(pruned_tree, setdiff(pruned_tree$tip.label, common_alpha))
alpha_pgls_df <- alpha_pgls_df[tree_alpha$tip.label, ] %>%
  mutate(ColonySize_log = log(ColonySize_U_mean + 1))

cat("Species in alpha diversity analysis:", nrow(alpha_pgls_df), "\n")
```

## 10.2 Phylogenetic signal: Blomberg's K and Pagel's λ

```{r alpha-phylosig}
shannon_vec <- setNames(alpha_pgls_df$Shannon_mean, rownames(alpha_pgls_df))

# Blomberg's K
k_shannon <- phylosig(tree_alpha, shannon_vec, method = "K", test = TRUE, nsim = 1000)

# Pagel's lambda
l_shannon <- phylosig(tree_alpha, shannon_vec, method = "lambda", test = TRUE)

cat("=== Blomberg's K ===\n")
cat(sprintf("Shannon — K = %.3f, p = %.4f\n", k_shannon$K, k_shannon$P))

cat("\n=== Pagel's lambda ===\n")
cat(sprintf("Shannon — lambda = %.3f, p = %.4f\n", l_shannon$lambda, l_shannon$P))
```

## 10.3 Ancestral state reconstruction (contMap)

```{r alpha-contmap, fig.width=8, fig.height=7}
contMap(tree_alpha, shannon_vec, fsize = c(0.5, 0.8), lwd = 3,
        outline = FALSE)
title("Shannon index mapped on phylogeny")
```

## 10.4 PGLS: Alpha diversity vs. colony size

```{r alpha-pgls}
spp_alpha   <- alpha_pgls_df$VoucherNumber
corBM_alpha <- corBrownian(phy = tree_alpha, form = ~spp_alpha)

pgls_Shannon_CS <- gls(Shannon_mean ~ ColonySize_log,
                        data = alpha_pgls_df, correlation = corBM_alpha)

summary(pgls_Shannon_CS)
```

## 10.5 Visualize PGLS: Alpha diversity vs. colony size

```{r alpha-pgls-plots, fig.width=8, fig.height=4}
make_alpha_pgls_plot <- function(mdl, y_col, y_label) {
  p_val  <- summary(mdl)$tTable[2, 4]
  r2     <- summary(lm(alpha_pgls_df[[y_col]] ~ alpha_pgls_df$ColonySize_log))$r.squared
  ggplot(alpha_pgls_df, aes(x = ColonySize_U_mean, y = .data[[y_col]])) +
    geom_point(shape = 21, fill = "grey50", size = 3, alpha = 0.8) +
    geom_smooth(method = "lm", se = FALSE, color = "grey30", linetype = "dashed") +
    scale_x_log10() +
    annotation_logticks(sides = "b") +
    annotate("text", x = Inf, y = Inf,
             label = paste0("R² = ", round(r2, 2), "\nPGLS p = ", signif(p_val, 3)),
             hjust = 1.1, vjust = 1.5, size = 3.5, fontface = "italic") +
    labs(x = "Colony Size (upper estimate)", y = y_label) +
    theme_bw() +
    theme(panel.grid.minor = element_blank(), panel.border = element_blank(),
          axis.line = element_line(colour = "black"),
          aspect.ratio = 1)
}

p_shannon <- make_alpha_pgls_plot(pgls_Shannon_CS, "Shannon_mean", "Shannon index")

p_shannon
ggsave("figures/raw/alpha_diversity_vs_colonySize.pdf", width = 4, height = 4)
```

---

# 11. Summary of Outputs

```{r summary}
cat("============================================\n")
cat("Files exported to data/processed/:\n")
cat("  - phyloseqOBJ.rds\n")
cat("  - alpha_table.rds\n")
cat("  - metadata_symbionts.rds\n")
cat("  - pruned_tree.rds\n")
cat("  - discrete_Sng.rds\n")
cat("  - discrete_Gill.rds\n")
cat("  - discrete_Lacto.rds\n")
cat("  - discrete_Comm.rds\n")
cat("  - unique_all_avg.rds\n")
cat("============================================\n")
cat("Figures exported to figures/raw/:\n")
cat("  - alpha_diversity.pdf\n")
cat("  - alpha_diversity_vs_colonySize.pdf\n")
cat("  - PGLS_relAbundance_colonySize.pdf\n")
cat("  - phyloANOVA_colonySize.pdf\n")
cat("============================================\n")
cat("These files are required inputs for:\n")
cat("  → 03_qPCR_and_AbsAbundance.Rmd\n")
```

---

# Session Info

```{r session-info}
sessionInfo()
```
