---
title: "03 - qPCR Analysis and Estimated Absolute Abundance"
author: "Nickole Villabona"
date: "2025"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: united
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
# Set working directory to project root so relative paths work when knitting from RStudio
knitr::opts_knit$set(root.dir = normalizePath(".."))

library(vroom)
library(tidyverse)
library(funrar)
library(phyloseq)
library(microViz)
library(ape)
library(phytools)
library(nlme)
library(cowplot)
library(gridExtra)
library(smplot2)

`%notin%` <- function(x, y) !(x %in% y)

# ── File paths ───────────────────────────────────────────────────────────────
path_raw  <- "data/raw/"
path_proc <- "data/processed/"

dir.create("figures/raw", showWarnings = FALSE, recursive = TRUE)
```

---

## Overview

This script processes quantitative PCR (qPCR) data measuring total bacterial load per
bee abdomen, and combines it with 16S rRNA relative abundance data to estimate
per-taxon absolute abundances for four focal gut symbionts (*Snodgrassella*,
*Gilliamella*, *Lactobacillus*, *Commensalibacter*). Phylogenetic comparative methods
(PGLS, phyloANOVA) are applied to test associations between symbiont absolute
abundance and host colony size.

**Inputs (from `data/processed/`):**
- `phyloseqOBJ.rds` — rarefied and filtered phyloseq object
- `metadata_symbionts.rds` — symbiont prevalence merged with metadata
- `pruned_tree.rds` — ultrametric tree pruned to study species
- `discrete_Sng.rds` — binary Snodgrassella presence/absence
- `discrete_Gill.rds` — binary Gilliamella presence/absence
- `unique_all_avg.rds` — species-averaged symbiont data
- `muestras_a_eliminar.rds` — samples removed due to high contaminant load

**Inputs (from `data/raw/`):**
- `qPCR_All_Triplicates_WB.csv` — raw qPCR triplicates
- `metadata_table_updated.tsv` — updated sample metadata

**Outputs (to `data/processed/`):**
- `qpcr_filtered.rds` — processed qPCR data (sample level)
- `estimated_abs_abundance_bySpp.tsv` — estimated absolute abundance per species
- `estimated_abs_abundance_byGenus.tsv` — estimated absolute abundance per bee genus

---

# 1. Load Outputs from Previous Scripts

```{r load-processed-data}
phyloseqOBJ        <- readRDS(paste0(path_proc, "phyloseqOBJ.rds"))
metadata_symbionts <- readRDS(paste0(path_proc, "metadata_symbionts.rds"))
pruned_tree        <- readRDS(paste0(path_proc, "pruned_tree.rds"))
discrete_Sng       <- readRDS(paste0(path_proc, "discrete_Sng.rds"))
discrete_Gill      <- readRDS(paste0(path_proc, "discrete_Gill.rds"))
unique_all_avg     <- readRDS(paste0(path_proc, "unique_all_avg.rds"))
muestras_a_eliminar <- readRDS(paste0(path_proc, "muestras_a_eliminar.rds"))

cat("All processed objects loaded successfully.\n")
```

---

# 2. qPCR Data Processing

## 2.1 Load and average triplicates

Each sample was run in triplicate. Here we average the three measurements and compute
copies per bee abdomen, accounting for the elution volume (20 µL) and the fraction of
homogenate extracted (400/750 µL).

```{r load-qpcr}
qPCR_Raw <- vroom(paste0(path_raw, "qPCR_All_Triplicates_WB.csv"), delim = ";") %>%
  mutate(Mean_Conc = as.numeric(gsub(",", ".", Mean_Conc)))

# Average triplicates and compute copies per bee abdomen
qPCR_Mean <- qPCR_Raw %>%
  mutate(SampleID = as.factor(SampleID)) %>%
  group_by(SampleID) %>%
  dplyr::summarise(Mean_Conc = mean(Mean_Conc, na.rm = TRUE)) %>%
  mutate(copies_perbee = Mean_Conc * 20 * (1 / (400 / 750)))

# Remove samples with very low DNA identified during library preparation
samples_to_remove_qpcr <- c("M001", "M013", "M025", "M037", "M049", "M06",
                             "M072", "M083", "M002", "M014", "M026", "M038",
                             "M050", "M062", "M073", "M084")

qpcr_filtered <- qPCR_Mean %>%
  filter(!SampleID %in% samples_to_remove_qpcr) %>%
  mutate(SampleID = as.character(SampleID))

cat("qPCR samples after filtering:", nrow(qpcr_filtered), "\n")
```

## 2.2 Merge with metadata

```{r merge-metadata}
metadata_raw <- vroom(paste0(path_raw, "metadata_table_updated.tsv")) %>%
  mutate(SampleID     = as.character(SampleID),
         CollectionID = as.character(CollectionID)) %>%
  distinct(SampleID, .keep_all = TRUE)

qpcr_merged <- qpcr_filtered %>%
  left_join(metadata_raw, by = "SampleID")

saveRDS(qpcr_filtered, file = paste0(path_proc, "qpcr_filtered.rds"))
```

## 2.3 Apply decontam sample quality flag

Samples identified in Script 01 as having >25% contaminant sequences are flagged here
for visualization. These samples are expected to have lower bacterial loads.

```{r decontam-flag}
qpcr_decontam <- qpcr_merged %>%
  mutate(
    SampleQuality = if_else(SampleID %in% muestras_a_eliminar, FALSE, TRUE),
    SampleType    = if_else(SampleID %in% c("MP1ExBl", "MP2ExBl"), "Blank", "Sample")
  )

blank_mean <- qpcr_decontam %>%
  filter(SampleType == "Blank") %>%
  summarise(mean_blank = mean(Mean_Conc, na.rm = TRUE)) %>%
  pull(mean_blank)
```

## 2.4 Visualize qPCR quality

Samples with high contaminant loads (FALSE) consistently show lower bacterial copy
numbers, often approaching extraction blank levels, validating the decontam filter.

```{r qpcr-quality-plot, fig.width=7, fig.height=6}
qpcr_decontam_plot <- ggplot(qpcr_decontam,
       aes(x = factor(SampleQuality), y = copies_perbee,
           fill = SampleQuality, color = SampleQuality)) +
  sm_raincloud(
    violin.params  = list(alpha = 0.7, width = 0.6),
    boxplot.params = list(width = 0.2, outlier.shape = NA, alpha = 0.7),
    point.params   = list(size = 3, alpha = 0.8)
  ) +
  geom_hline(yintercept = blank_mean, linetype = "dashed",
             colour = "blue", linewidth = 0.3) +
  annotate("text", x = Inf, y = blank_mean, label = "Blanks mean",
           hjust = 1.05, vjust = -0.4, size = 4) +
  scale_y_log10() +
  scale_fill_manual(values  = sm_color("lightpurple", "green")) +
  scale_color_manual(values = sm_color("lightpurple", "green")) +
  labs(x = "Sample Quality", y = "16S rRNA gene copies",
       title = "qPCR: Decontam filter") +
  theme_minimal(base_size = 14)

qpcr_decontam_plot
ggsave("figures/raw/qPCR_decontam_filter.pdf", width = 7, height = 6)
```

---

# 3. Extract Symbiont Relative Abundances

```{r extract-rel-abundances}
# Aggregate phyloseq to genus level
ps.genus <- phyloseqOBJ %>% tax_agg(rank = "Genus")

Prevalence_Genus <- as.data.frame(make_relative(t(ps.genus@otu_table)))

# Extract four focal symbionts at genus level
Prev_Snodgrassella    <- Prevalence_Genus %>% select(Snodgrassella)    %>% rownames_to_column("SampleID")
Prev_Gilliamella      <- Prevalence_Genus %>% select(Gilliamella)      %>% rownames_to_column("SampleID")
Prev_Lactobacillus    <- Prevalence_Genus %>% select(Lactobacillus)    %>% rownames_to_column("SampleID")
Prev_Commensalibacter <- Prevalence_Genus %>% select(Commensalibacter) %>% rownames_to_column("SampleID")

Prev_Snod_Gill <- Prev_Snodgrassella %>%
  left_join(Prev_Gilliamella,      by = "SampleID") %>%
  left_join(Prev_Lactobacillus,    by = "SampleID") %>%
  left_join(Prev_Commensalibacter, by = "SampleID")

cat("Relative abundance table — first rows:\n")
head(Prev_Snod_Gill)
```

---

# 4. Estimated Absolute Abundance

The estimated per-taxon absolute abundance is computed by multiplying the relative
proportion of each symbiont by the total 16S rRNA gene copies per bee (from qPCR).
This approach provides a rough estimate of absolute symbiont abundance while
accounting for differences in total bacterial load across species.

```{r estimated-absolute-abundance}
# Join relative abundances with qPCR total copies at the sample level
estimated_abs <- Prev_Snod_Gill %>%
  left_join(qpcr_filtered[, c("SampleID", "copies_perbee")], by = "SampleID") %>%
  mutate(
    Snod_abs_estimated  = Snodgrassella    * copies_perbee,
    Gill_abs_estimated  = Gilliamella      * copies_perbee,
    Lacto_abs_estimated = Lactobacillus    * copies_perbee,
    Comm_abs_estimated  = Commensalibacter * copies_perbee
  )

cat("Samples with both 16S and qPCR data:",
    sum(!is.na(estimated_abs$copies_perbee)), "\n")
cat("Samples missing qPCR data:",
    sum(is.na(estimated_abs$copies_perbee)), "\n")

head(estimated_abs)
```

## 4.1 Average estimated absolute abundances by species

Estimated absolute abundances are averaged per voucher (species) to align with the
phylogenetic tree for comparative analyses.

```{r average-by-species}
# Build VoucherNumber mapping from tree tips
df_tree_ult1 <- data.frame(VoucherNumber = pruned_tree$tip.label) %>%
  mutate(CollectionID = str_extract(VoucherNumber, "\\d+"))

# Merge estimated abundances with metadata and voucher mapping
metadata_est <- metadata_symbionts %>%
  select(SampleID, CollectionID, Genus, Species,
         ColonySize_U, BodySize) %>%
  distinct()

estimated_abs_full <- estimated_abs %>%
  left_join(metadata_est %>%
              mutate(CollectionID = as.character(CollectionID)),
            by = "SampleID") %>%
  left_join(df_tree_ult1, by = "CollectionID") %>%
  filter(!is.na(VoucherNumber))

# Average per VoucherNumber (species level)
estimated_abs_bySpp <- estimated_abs_full %>%
  group_by(VoucherNumber) %>%
  dplyr::summarise(
    Snod_abs_mean      = mean(Snod_abs_estimated,  na.rm = TRUE),
    Gill_abs_mean      = mean(Gill_abs_estimated,  na.rm = TRUE),
    Lacto_abs_mean     = mean(Lacto_abs_estimated, na.rm = TRUE),
    Comm_abs_mean      = mean(Comm_abs_estimated,  na.rm = TRUE),
    Snod_rel_mean      = mean(Snodgrassella,        na.rm = TRUE),
    Gill_rel_mean      = mean(Gilliamella,           na.rm = TRUE),
    Lacto_rel_mean     = mean(Lactobacillus,         na.rm = TRUE),
    Comm_rel_mean      = mean(Commensalibacter,      na.rm = TRUE),
    copies_perbee_mean = mean(copies_perbee,         na.rm = TRUE),
    copies_perbee_sem  = sd(copies_perbee,           na.rm = TRUE),
    ColonySize_U_mean  = mean(ColonySize_U,          na.rm = TRUE),
    BodySize_mean      = mean(BodySize,               na.rm = TRUE),
    Bee_Genus          = first(Genus),
    n_samples          = n(),
    .groups = "drop"
  ) %>%
  as.data.frame()

rownames(estimated_abs_bySpp) <- estimated_abs_bySpp$VoucherNumber

cat("Species with estimated absolute abundances:", nrow(estimated_abs_bySpp), "\n")

# Export
write.table(estimated_abs_bySpp,
            file      = paste0(path_proc, "estimated_abs_abundance_bySpp.tsv"),
            sep       = "\t", row.names = FALSE, quote = FALSE)
```

---

# 5. Visualizations

## 5.1 Barplots: estimated absolute abundance by bee genus

```{r barplots-by-genus, fig.width=14, fig.height=6}
symbiont_colors <- c("Snodgrassella"    = "#e54b4b",
                     "Gilliamella"      = "#2b9dce",
                     "Lactobacillus"    = "#6ab187",
                     "Commensalibacter" = "#f0a500")

plot_genus_df <- estimated_abs_bySpp %>%
  filter(!is.na(Bee_Genus)) %>%
  group_by(Bee_Genus) %>%
  dplyr::summarise(
    Snod_mean  = mean(Snod_abs_mean,  na.rm = TRUE),
    Gill_mean  = mean(Gill_abs_mean,  na.rm = TRUE),
    Lacto_mean = mean(Lacto_abs_mean, na.rm = TRUE),
    Comm_mean  = mean(Comm_abs_mean,  na.rm = TRUE),
    Snod_se    = sd(Snod_abs_mean,  na.rm = TRUE) / sqrt(n()),
    Gill_se    = sd(Gill_abs_mean,  na.rm = TRUE) / sqrt(n()),
    Lacto_se   = sd(Lacto_abs_mean, na.rm = TRUE) / sqrt(n()),
    Comm_se    = sd(Comm_abs_mean,  na.rm = TRUE) / sqrt(n()),
    n          = n(), .groups = "drop"
  ) %>%
  pivot_longer(
    cols      = c(Snod_mean, Gill_mean, Lacto_mean, Comm_mean),
    names_to  = "Symbiont",
    values_to = "AbsAbundance"
  ) %>%
  mutate(
    SE = case_when(
      Symbiont == "Snod_mean"  ~ Snod_se,
      Symbiont == "Gill_mean"  ~ Gill_se,
      Symbiont == "Lacto_mean" ~ Lacto_se,
      Symbiont == "Comm_mean"  ~ Comm_se
    ),
    Symbiont = case_when(
      Symbiont == "Snod_mean"  ~ "Snodgrassella",
      Symbiont == "Gill_mean"  ~ "Gilliamella",
      Symbiont == "Lacto_mean" ~ "Lactobacillus",
      Symbiont == "Comm_mean"  ~ "Commensalibacter"
    )
  )

genus_order <- plot_genus_df %>%
  filter(Symbiont == "Snodgrassella") %>%
  arrange(desc(AbsAbundance)) %>%
  pull(Bee_Genus)

plot_genus_df$Bee_Genus <- factor(plot_genus_df$Bee_Genus, levels = genus_order)
plot_genus_df$Symbiont  <- factor(plot_genus_df$Symbiont,
                                   levels = names(symbiont_colors))

ggplot(plot_genus_df, aes(x = Bee_Genus, y = AbsAbundance, fill = Symbiont)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8),
           width = 0.7, alpha = 0.85) +
  geom_errorbar(aes(ymin = AbsAbundance - SE, ymax = AbsAbundance + SE),
                position = position_dodge(width = 0.8),
                width = 0.25, color = "grey40") +
  scale_fill_manual(values = symbiont_colors,
                    labels = ~ parse(text = paste0("italic('", ., "')"))) +
  scale_y_continuous(trans = "log1p", labels = scales::label_scientific()) +
  labs(x = "Bee Genus", y = "Estimated copies per bee (log scale)", fill = NULL) +
  theme_bw(base_size = 12) +
  theme(axis.text.x       = element_text(angle = 45, hjust = 1, face = "italic"),
        panel.grid.major.x = element_blank(),
        panel.border       = element_blank(),
        axis.line          = element_line(colour = "black"),
        legend.position    = "top",
        legend.text        = element_text(face = "italic"))

ggsave("figures/raw/barplot_absAbundance_byGenus.pdf", width = 14, height = 6)

write.table(plot_genus_df,
            file = paste0(path_proc, "estimated_abs_abundance_byGenus.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)
```

## 5.2 Boxplots: estimated absolute abundance by bee genus

```{r boxplots-by-genus, fig.width=14, fig.height=8}
symbiont_colors <- c("Snodgrassella"    = "#e54b4b",
                     "Gilliamella"      = "#2b9dce",
                     "Lactobacillus"    = "#6ab187",
                     "Commensalibacter" = "#f0a500")

# Species-level data in long format (one row per species per symbiont)
box_df <- estimated_abs_bySpp %>%
  filter(!is.na(Bee_Genus)) %>%
  pivot_longer(cols      = c(Snod_abs_mean, Gill_abs_mean,
                              Lacto_abs_mean, Comm_abs_mean),
               names_to  = "Symbiont",
               values_to = "AbsAbundance") %>%
  mutate(Symbiont = recode(Symbiont,
           "Snod_abs_mean"  = "Snodgrassella",
           "Gill_abs_mean"  = "Gilliamella",
           "Lacto_abs_mean" = "Lactobacillus",
           "Comm_abs_mean"  = "Commensalibacter"),
         Symbiont = factor(Symbiont, levels = names(symbiont_colors)))

# Order bee genera by median Snodgrassella absolute abundance
genus_order_box <- box_df %>%
  filter(Symbiont == "Snodgrassella") %>%
  group_by(Bee_Genus) %>%
  summarise(med = median(AbsAbundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(med)) %>%
  pull(Bee_Genus)

box_df$Bee_Genus <- factor(box_df$Bee_Genus, levels = genus_order_box)

ggplot(box_df, aes(x = Bee_Genus, y = pmax(AbsAbundance, 1e-04),
                    fill = Symbiont, color = Symbiont)) +
  geom_boxplot(alpha = 0.35, outlier.shape = NA, width = 0.6) +
  geom_jitter(width = 0.15, size = 0.7, alpha = 0.5) +
  scale_y_log10(labels = scales::label_scientific()) +
  scale_fill_manual(values  = symbiont_colors) +
  scale_color_manual(values = symbiont_colors) +
  facet_wrap(~ Symbiont, nrow = 2, scales = "free_y") +
  labs(x = NULL, y = "Estimated copies/bee (log scale)") +
  theme_bw(base_size = 11) +
  theme(axis.text.x      = element_text(angle = 45, hjust = 1, face = "italic"),
        panel.grid.minor = element_blank(),
        strip.text       = element_text(face = "italic"),
        legend.position  = "none")

ggsave("figures/raw/boxplot_absAbundance_byGenus_panels.pdf", width = 14, height = 8)
```

---

# 6. Boxplots: Bacterial Load vs. Symbiont Presence/Absence

```{r qpcr-presence-plots, fig.width=8, fig.height=4}
# Merge qPCR species averages with symbiont presence/absence
qpcr_merged_bySpp <- estimated_abs_bySpp %>%
  select(VoucherNumber, copies_perbee_mean, copies_perbee_sem,
         ColonySize_U_mean, BodySize_mean, Bee_Genus)

qpcr_bySpp_Snod_Gill <- qpcr_merged_bySpp %>%
  left_join(discrete_Sng, by = "VoucherNumber") %>%
  left_join(discrete_Gill, by = "VoucherNumber") %>%
  mutate(
    Snodgrassella_pre = factor(Discrete_Snodgrassella,
                               levels = c(0, 1), labels = c("abs", "pres")),
    Gilliamella_pre   = factor(Discrete_Gilliamella,
                               levels = c(0, 1), labels = c("abs", "pres"))
  )

plot_snod <- ggplot(qpcr_bySpp_Snod_Gill,
                    aes(x = Snodgrassella_pre, y = log10(copies_perbee_mean))) +
  geom_boxplot(aes(fill = Snodgrassella_pre), colour = "grey40",
               width = 0.6, outlier.shape = NA, alpha = 0.5) +
  geom_jitter(aes(color = Snodgrassella_pre), width = 0.1, size = 2.3, alpha = 0.8) +
  scale_fill_manual(values  = c("abs" = "white",  "pres" = "#f4a3a3")) +
  scale_color_manual(values = c("abs" = "grey60", "pres" = "#e54b4b")) +
  labs(x = expression(italic("Snodgrassella") ~ "presence"),
       y = "16S rRNA gene copies (log10)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

plot_gill <- ggplot(qpcr_bySpp_Snod_Gill,
                    aes(x = Gilliamella_pre, y = log10(copies_perbee_mean))) +
  geom_boxplot(aes(fill = Gilliamella_pre), colour = "grey40",
               width = 0.6, outlier.shape = NA, alpha = 0.5) +
  geom_jitter(aes(color = Gilliamella_pre), width = 0.1, size = 2.3, alpha = 0.8) +
  scale_fill_manual(values  = c("abs" = "white",  "pres" = "#a9d2f3")) +
  scale_color_manual(values = c("abs" = "grey60", "pres" = "#2b9dce")) +
  labs(x = expression(italic("Gilliamella") ~ "presence"),
       y = "16S rRNA gene copies (log10)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

cowplot::plot_grid(plot_snod, plot_gill, labels = c("A", "B"),
                   nrow = 1, align = "h")
ggsave("figures/raw/qPCR_symbiont_presence.pdf", width = 8, height = 4)
```

---

# 7. Phylogenetic ANOVA: Bacterial Load vs. Symbiont Presence

```{r qpcr-phylo-anova}
# Prepare species-level table
df_qpcr <- qpcr_merged_bySpp[, c("VoucherNumber", "copies_perbee_mean")] %>%
  mutate(log10_load = log10(copies_perbee_mean))
rownames(df_qpcr) <- df_qpcr$VoucherNumber

# Merge with presence/absence data
pres_tbl <- qpcr_bySpp_Snod_Gill[, c("VoucherNumber",
                                       "Snodgrassella_pre", "Gilliamella_pre")]
rownames(pres_tbl) <- pres_tbl$VoucherNumber

dat_pa <- cbind(
  df_qpcr[rownames(pres_tbl), "log10_load", drop = FALSE],
  pres_tbl[, c("Snodgrassella_pre", "Gilliamella_pre")]
)

# Drop species with missing qPCR data, then align with tree
dat_pa <- dat_pa[!is.na(dat_pa$log10_load), ]

common_pa <- intersect(pruned_tree$tip.label, rownames(dat_pa))
tree_pa   <- drop.tip(pruned_tree, setdiff(pruned_tree$tip.label, common_pa))
dat_pa    <- dat_pa[tree_pa$tip.label, ]

cat("Species in phylo-ANOVA:", nrow(dat_pa), "\n")

y_load <- setNames(dat_pa$log10_load,         rownames(dat_pa))
x_snod <- setNames(dat_pa$Snodgrassella_pre,  rownames(dat_pa))
x_gill <- setNames(dat_pa$Gilliamella_pre,    rownames(dat_pa))

res_snod_qpcr <- phylANOVA(tree_pa, x = x_snod, y = y_load, nsim = 1000)
res_gill_qpcr <- phylANOVA(tree_pa, x = x_gill, y = y_load, nsim = 1000)

cat(sprintf("Snodgrassella: F = %.2f, p = %.3f\n",
            unname(res_snod_qpcr$F), unname(res_snod_qpcr$Pf)))
cat(sprintf("Gilliamella:   F = %.2f, p = %.3f\n",
            unname(res_gill_qpcr$F), unname(res_gill_qpcr$Pf)))
```

---

# 8. PGLS: Estimated Absolute Abundance vs. Colony Size

```{r pgls-setup}
# Align data with tree
common_pgls      <- intersect(pruned_tree$tip.label,
                               estimated_abs_bySpp$VoucherNumber)
pruned_tree_pgls <- drop.tip(pruned_tree,
                              setdiff(pruned_tree$tip.label, common_pgls))

pgls_df <- estimated_abs_bySpp[common_pgls, ] %>%
  mutate(
    Snod_abs_log   = log(Snod_abs_mean   + 1),
    Gill_abs_log   = log(Gill_abs_mean   + 1),
    Lacto_abs_log  = log(Lacto_abs_mean  + 1),
    Comm_abs_log   = log(Comm_abs_mean   + 1),
    ColonySize_log = log(ColonySize_U_mean + 1)
  ) %>%
  filter(!is.na(ColonySize_log), !is.na(VoucherNumber),
         !is.na(Snod_abs_log), !is.na(Gill_abs_log),
         !is.na(Lacto_abs_log), !is.na(Comm_abs_log))

# Re-prune tree to match cleaned data
pruned_tree_pgls <- drop.tip(pruned_tree_pgls,
                              setdiff(pruned_tree_pgls$tip.label,
                                      pgls_df$VoucherNumber))
pgls_df <- pgls_df[pruned_tree_pgls$tip.label, ]

cat("Rows in PGLS table:", nrow(pgls_df), "\n")
cat("Tips in tree:", length(pruned_tree_pgls$tip.label), "\n")

spp   <- pgls_df$VoucherNumber
corBM <- corBrownian(phy = pruned_tree_pgls, form = ~spp)
```

```{r pgls-models}
pgls_Snod_abs_CS  <- gls(Snod_abs_log  ~ ColonySize_log, data = pgls_df, correlation = corBM)
pgls_Gill_abs_CS  <- gls(Gill_abs_log  ~ ColonySize_log, data = pgls_df, correlation = corBM)
pgls_Lacto_abs_CS <- gls(Lacto_abs_log ~ ColonySize_log, data = pgls_df, correlation = corBM)
pgls_Comm_abs_CS  <- gls(Comm_abs_log  ~ ColonySize_log, data = pgls_df, correlation = corBM)

cat("=== PGLS: estimated absolute abundance vs. Colony Size ===\n")
for (lst in list(
  list(pgls_Snod_abs_CS,  "Snodgrassella"),
  list(pgls_Gill_abs_CS,  "Gilliamella"),
  list(pgls_Lacto_abs_CS, "Lactobacillus"),
  list(pgls_Comm_abs_CS,  "Commensalibacter")
)) {
  cat(sprintf("%-20s p = %.4f  slope = %+.3f\n",
              lst[[2]], summary(lst[[1]])$tTable[2, 4], summary(lst[[1]])$tTable[2, 1]))
}
```

## 8.1 Visualize PGLS results

```{r pgls-plots, fig.width=10, fig.height=10}
# Common axis limits across all four panels for direct comparability
x_lims_pgls <- range(pgls_df$ColonySize_U_mean, na.rm = TRUE) * c(0.5, 2)
y_vals_pgls  <- c(pmax(pgls_df$Snod_abs_mean,  1e-04),
                  pmax(pgls_df$Gill_abs_mean,  1e-04),
                  pmax(pgls_df$Lacto_abs_mean, 1e-04),
                  pmax(pgls_df$Comm_abs_mean,  1e-04))
y_lims_pgls  <- range(y_vals_pgls, na.rm = TRUE) * c(0.5, 2)

make_pgls_plot <- function(mdl, y_col, fill_col, label) {
  p_val  <- summary(mdl)$tTable[2, 4]
  sub_df <- pgls_df[pgls_df[[y_col]] > 1e-03, ]
  r2     <- summary(lm(log(sub_df[[y_col]]) ~ log(sub_df$ColonySize_U_mean)))$r.squared
  ggplot(pgls_df, aes(x = ColonySize_U_mean, y = pmax(.data[[y_col]], 1e-04))) +
    geom_point(shape = 21, fill = fill_col, size = 3, alpha = 0.8) +
    geom_smooth(method = "lm", se = FALSE, color = "grey", linetype = "dashed") +
    scale_x_log10(limits = x_lims_pgls) +
    scale_y_log10(limits = y_lims_pgls) +
    annotation_logticks() +
    annotate("text", x = Inf, y = Inf,
             label = paste0("R² = ", round(r2, 2), "\nPGLS p = ", signif(p_val, 3)),
             hjust = 1.1, vjust = 1.5, size = 3.5, fontface = "italic") +
    labs(x = "Colony Size",
         y = bquote(italic(.(label)) ~ "estimated copies/bee")) +
    theme_bw() +
    theme(panel.grid.minor = element_blank(), panel.border = element_blank(),
          axis.line = element_line(colour = "black"),
          aspect.ratio = 1)
}

p1 <- make_pgls_plot(pgls_Snod_abs_CS,  "Snod_abs_mean",  "#e54b4b", "Snodgrassella")
p2 <- make_pgls_plot(pgls_Gill_abs_CS,  "Gill_abs_mean",  "#2b9dce", "Gilliamella")
p3 <- make_pgls_plot(pgls_Lacto_abs_CS, "Lacto_abs_mean", "#6ab187", "Lactobacillus")
p4 <- make_pgls_plot(pgls_Comm_abs_CS,  "Comm_abs_mean",  "#f0a500", "Commensalibacter")

cowplot::plot_grid(p1, p2, p3, p4, labels = c("A", "B", "C", "D"), nrow = 2)
ggsave("figures/raw/PGLS_absAbundance_colonySize.pdf", width = 10, height = 10)
```

---

# 9. Summary of Outputs

```{r summary}
cat("============================================\n")
cat("Files exported to data/processed/:\n")
cat("  - qpcr_filtered.rds\n")
cat("  - estimated_abs_abundance_bySpp.tsv\n")
cat("  - estimated_abs_abundance_byGenus.tsv\n")
cat("Figures exported to figures/raw/:\n")
cat("  - qPCR_decontam_filter.pdf\n")
cat("  - qPCR_symbiont_presence.pdf\n")
cat("  - barplot_absAbundance_byGenus.pdf\n")
cat("  - barplot_absAbundance_byGenus_panels.pdf\n")
cat("  - PGLS_absAbundance_colonySize.pdf\n")
cat("============================================\n")
```

---

# Session Info

```{r session-info}
sessionInfo()
```
