---
title: "Costa Rica Dataset - Stingless Bee Gut Microbiome"
author: "Nickole Villabona"
date: "2025"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
# Set working directory to project root so relative paths work when knitting from RStudio
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r libraries}
library(phyloseq)
library(microViz)
library(vegan)
library(ape)
library(tidyverse)
library(RColorBrewer)
library(cowplot)
library(ggtext)
library(patchwork)
library(wesanderson)
```

```{r species-colors}
# Master color palette — used in PCoA, PCA, rarefaction curves, and map
sp_colors_org <- c(
  "Partamona_orizabaensis"         = "#FC4E07",
  "Scaptotrigona_subobscuripennis" = "#1B7B3D",
  "Tetragonisca_angustula"         = "#00AFBB",
  "Tetragona_ziegleri"            = "#9d4edd",
  "Trigona_fulviventris"           = "#E7B800"
)
```

---

# 1. Data loading

```{r load-data}
path_raw <- "data/raw/"
dir.create("figures/raw", showWarnings = FALSE, recursive = TRUE)

# ASV counts (exported from the QIIME 2 artifact; 1,928 ASVs x 191 samples)
counts_jb <- read.delim(paste0(path_raw, "CostaRica_ASV_table.tsv"),
                        row.names = 1, check.names = FALSE) %>% as.matrix()

# Specimen records. Organism and ColonyName are rebuilt here under the names the
# rest of this script expects; the five extraction blanks have no specimen record
# and so get NA Organism, which section 4 drops.
metadata_jb <- read.delim(paste0(path_raw, "CostaRica_metadata.tsv"),
                          stringsAsFactors = FALSE) %>%
  filter(!is.na(SampleID)) %>%
  mutate(Organism   = paste(Genus, Species, sep = "_"),
         ColonyName = ColonyID)
rownames(metadata_jb) <- metadata_jb$SampleID
metadata_jb <- metadata_jb[colnames(counts_jb), , drop = FALSE]
rownames(metadata_jb) <- colnames(counts_jb)

taxonomy_jb <- read_tsv(paste0(path_raw, "CostaRica_taxonomy.tsv"),
                        show_col_types = FALSE,
                        col_types = cols(.default = col_character())) %>%
  dplyr::rename(FeatureID = ASV_ID)

obj_joyce <- phyloseq(otu_table(counts_jb, taxa_are_rows = TRUE),
                      sample_data(metadata_jb))
obj_joyce
```

---

# 2. Taxonomy cleaning

Parse SILVA taxonomy string and replace with cleaned table.

```{r taxonomy-parse}
tax_df_jb <- taxonomy_jb %>%
  separate_wider_delim(
    Taxon, delim = ";",
    names = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"),
    too_few = "align_start"
  ) %>%
  mutate(across(Domain:Species, ~ sub("^[a-zA-Z]__", "", trimws(.x)))) %>%
  mutate(FeatureID = trimws(FeatureID)) %>%
  distinct(FeatureID, .keep_all = TRUE) %>%
  as.data.frame()

rownames(tax_df_jb) <- tax_df_jb$FeatureID

# Rename Domain → Kingdom and ensure all ranks exist
desired <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
tax_df_jb$Kingdom <- tax_df_jb$Domain
tax_df_jb$Domain  <- NULL
for (col in setdiff(desired, colnames(tax_df_jb))) tax_df_jb[[col]] <- NA_character_
tax_df_jb <- tax_df_jb[, desired, drop = FALSE]
```

```{r taxonomy-update}
ids <- taxa_names(obj_joyce)

missing_asvs <- setdiff(ids, rownames(tax_df_jb))
if (length(missing_asvs)) {
  # Merge with old taxonomy for ASVs not in new table
  old_tax <- as.data.frame(tax_table(obj_joyce), stringsAsFactors = FALSE)
  for (col in setdiff(colnames(tax_df_jb), colnames(old_tax))) old_tax[[col]] <- NA_character_
  for (col in setdiff(colnames(old_tax), colnames(tax_df_jb))) tax_df_jb[[col]] <- NA_character_
  comunes <- intersect(rownames(old_tax), rownames(tax_df_jb))
  old_tax[comunes, colnames(tax_df_jb)] <- tax_df_jb[comunes, colnames(tax_df_jb)]
  new_mat <- as.matrix(old_tax[ids, , drop = FALSE])
} else {
  new_mat <- as.matrix(tax_df_jb[ids, , drop = FALSE])
}
storage.mode(new_mat) <- "character"
tax_table(obj_joyce)  <- tax_table(new_mat)

rank_names(obj_joyce)
```

Manual LCA corrections confirmed by BLAST (≥95% identity), plus unification of
the SILVA typo `Giliamella` (one 'l') → `Gilliamella`:

```{r taxonomy-lca-fix}
tax <- tax_table(obj_joyce)

# ASV-level corrections
lca_corrections <- c(
  "2f5bd1cf50b6d70e150bc1829af4ee7e" = "Gilliamella",
  "9c489be807c8d254d8af389af7b2ad60" = "Snodgrassella",
  "c1d35f5da6018f72638c83306839f516" = "Gilliamella",
  "07dc3714ead232d81471f5fa6f65f4a8" = "Gilliamella"
)

for (asv in names(lca_corrections)) {
  if (asv %in% rownames(tax)) {
    tax[asv, "Genus"] <- lca_corrections[asv]
  }
}

# Fix SILVA typo: "Giliamella" (1 'l') → "Gilliamella"
tax[tax[, "Genus"] %in% "Giliamella", "Genus"] <- "Gilliamella"

tax_table(obj_joyce) <- tax
cat("ASVs reclassified as Gilliamella (typo fix):",
    sum(tax[, "Genus"] == "Gilliamella", na.rm = TRUE), "\n")
```

---

# 3. Rarefaction curves

```{r rarefaction-curves}
otu_joyce <- as(otu_table(obj_joyce), "matrix")
if (taxa_are_rows(obj_joyce)) otu_joyce <- t(otu_joyce)

rare_list <- rarecurve(otu_joyce, step = 100,
                       sample = min(rowSums(otu_joyce)), label = FALSE)

rare_df <- map_dfr(seq_along(rare_list), function(i) {
  tibble(
    SampleID         = rownames(otu_joyce)[i],
    Depth            = attr(rare_list[[i]], "Subsample"),
    Rarefied_Richness = rare_list[[i]]
  )
}) %>%
  left_join(metadata_jb, by = "SampleID")

ggplot(rare_df, aes(x = Depth, y = Rarefied_Richness, group = SampleID, color = SampleID)) +
  geom_line(alpha = 0.8) +
  geom_vline(xintercept = 1000, color = "darkorange", linetype = "dashed") +
  coord_cartesian(xlim = c(0, 1500)) +
  facet_wrap(~ Organism) +
  labs(x = "Number of reads", y = "Observed richness",
       title = "Rarefaction curves per sample") +
  theme_minimal() +
  theme(legend.position = "none")
```

---

# 4. Filtering and normalization

Rarefy to 2,500 reads per sample; remove samples with missing `Organism`.

```{r filter-rarefy}
set.seed(123)
physeq_rarefied <- rarefy_even_depth(
  obj_joyce, sample.size = 2500,
  rngseed = 123, replace = FALSE, trimOTUs = TRUE, verbose = TRUE
)

physeq_clean <- physeq_rarefied %>%
  subset_samples(!is.na(Organism)) %>%
  tax_fix() %>%
  phyloseq_validate() %>%
  tax_fix(unknowns = c("uncultured"))

physeq_clean
```

---

# 5. Community composition

## 5.1 Iris plot (CLR-PCA)

```{r iris-plot}
physeq_clean %>%
  tax_filter(min_prevalence = 2.5 / 100, verbose = FALSE) %>%
  tax_transform(rank = "Genus", trans = "clr", zero_replace = "halfmin") %>%
  ord_calc(method = "PCA") %>%
  ord_plot_iris(
    tax_level = "Genus", n_taxa = 10, other = "Other",
    anno_colour = "Organism",
    anno_colour_style = list(alpha = 0.6, size = 3, show.legend = TRUE)
  )
```

## 5.2 Composition barplot by species

```{r comp-barplot-genus}
physeq_clean %>%
  comp_barplot(
    tax_level = "Genus", n_taxa = 15, other_name = "Other",
    taxon_renamer = function(x) stringr::str_remove(x, " [ae]t rel."),
    palette = distinct_palette(n = 15, add = "grey90"),
    merge_other = FALSE, bar_outline_colour = "darkgrey"
  ) +
  coord_flip() +
  facet_wrap("Organism", nrow = 1, scales = "free") +
  labs(x = NULL, y = NULL) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

```{r comp-barplot-family}
physeq_clean %>%
  comp_barplot(
    tax_level = "Family", n_taxa = 15, other_name = "Other",
    taxon_renamer = function(x) stringr::str_remove(x, " [ae]t rel."),
    palette = distinct_palette(n = 15, add = "grey90"),
    merge_other = FALSE, bar_outline_colour = "darkgrey"
  ) +
  coord_flip() +
  facet_wrap("Organism", nrow = 1, scales = "free") +
  labs(x = NULL, y = NULL) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

---

# 6. Beta diversity

```{r clr-transform}
physeq_clr <- physeq_clean %>%
  tax_transform(trans = "clr", rank = "Genus")
```

## 6.1 PCoA (Bray-Curtis)

```{r pcoa}
physeq_pcoa <- physeq_clean %>%
  tax_transform(trans = "identity") %>%
  dist_calc(dist = "bray") %>%
  ord_calc(method = "PCoA")

ord_plot(physeq_pcoa, color = "Organism", size = 4, alpha = 0.7) +
  scale_color_manual(values = sp_colors_org, name = "Bee species") +
  theme_classic() +
  labs(color = "Bee species")
```

## 6.2 PCA (CLR) — all species

```{r pca-all}
physeq_pca <- physeq_clr %>%
  ord_calc(method = "PCA")

range_val <- physeq_pca %>%
  ord_get() %>%
  vegan::scores(display = "sites") %>%
  .[, 1:2] %>% abs() %>% max()

p_pca_all <- ord_plot(physeq_pca, colour = "Organism",
                      size = 5, alpha = 0.7) +
  scale_color_manual(values = sp_colors_org, name = "Bee species") +
  theme_classic(base_size = 11) +
  labs(color = "Bee species") +
  coord_fixed()

p_pca_all
```

## 6.3 PCA per species (by colony)

```{r pca-per-species, fig.width=12, fig.height=6}
physeq_angustula    <- subset_samples(physeq_clr, Organism == "Tetragonisca_angustula")
physeq_orizabaensis <- subset_samples(physeq_clr, Organism == "Partamona_orizabaensis")

pca_angustula    <- ord_calc(physeq_angustula,    method = "PCA")
pca_orizabaensis <- ord_calc(physeq_orizabaensis, method = "PCA")

p_angustula <- ord_plot(pca_angustula, colour = "ColonyName",
                        size = 3.5, alpha = 0.7) +
  scale_color_manual(values = wes_palette(n = 4, name = "AsteroidCity3")) +
  theme_classic(base_size = 10) +
  ggtitle(expression(italic("Tetragonisca angustula"))) +
  labs(color = "Colony")

p_orizabaensis <- ord_plot(pca_orizabaensis, colour = "ColonyName",
                           size = 3.5, alpha = 0.7) +
  scale_color_manual(values = c("#FC4E07", "#ffa987")) +
  theme_classic(base_size = 10) +
  ggtitle(expression(italic("Partamona orizabaensis"))) +
  labs(color = "Colony")

right_col <- plot_grid(p_angustula, p_orizabaensis,
                       ncol = 1, align = "hv", axis = "tblr",
                       labels = c("B", "C"), label_size = 14)

fig_s2 <- plot_grid(p_pca_all + theme(legend.position = "right"),
                    right_col,
                    labels = c("A", ""), rel_widths = c(1.6, 1), align = "h")
fig_s2

ggsave("figures/raw/FigS2_ordination.pdf",
       plot = fig_s2, device = "pdf", width = 32, height = 18, units = "cm")
```

## 6.4 PCoA (Bray-Curtis) — *T. angustula*, colored by Gilliamella presence/absence

```{r pcoa-angustula-gilliamella, fig.width=6, fig.height=5}
# Subset rarefied (non-CLR) data to T. angustula
physeq_ang_raw <- subset_samples(physeq_clean, Organism == "Tetragonisca_angustula")

# Gilliamella relative abundance per sample
ps_rel_ang  <- transform_sample_counts(physeq_ang_raw, function(x) x / sum(x))
ps_gill_ang <- ps_rel_ang %>%
  subset_taxa(Genus == "Gilliamella") %>%
  tax_glom(taxrank = "Genus")

gill_df <- psmelt(ps_gill_ang) %>%
  select(Sample, Abundance) %>%
  rename(SampleID = Sample, gill_abun = Abundance)

# Add Gilliamella P/A (cutoff: >1% per individual) to sample_data
meta_ang <- data.frame(sample_data(physeq_ang_raw), stringsAsFactors = FALSE)
meta_ang$SampleID <- rownames(meta_ang)
meta_ang <- meta_ang %>%
  left_join(gill_df, by = "SampleID") %>%
  mutate(gill_abun      = replace_na(gill_abun, 0),
         Gilliamella_PA = ifelse(gill_abun > 0.01, "Present", "Absent"))
rownames(meta_ang) <- meta_ang$SampleID
sample_data(physeq_ang_raw) <- sample_data(meta_ang[sample_names(physeq_ang_raw), ])

# PCoA Bray-Curtis
pcoa_ang_bray <- physeq_ang_raw %>%
  tax_transform(trans = "identity") %>%
  dist_calc(dist = "bray") %>%
  ord_calc(method = "PCoA")

# PERMANOVA
otu_ang   <- as.data.frame(t(otu_table(physeq_ang_raw)))
meta_perm <- meta_ang[rownames(otu_ang), ]
perm_bray <- adonis2(otu_ang ~ Gilliamella_PA,
                     data = meta_perm, method = "bray",
                     permutations = 999, by = "margin")

r2   <- round(perm_bray["Gilliamella_PA", "R2"], 3)
pv   <- perm_bray["Gilliamella_PA", "Pr(>F)"]
plab <- ifelse(pv < 0.001, "p < 0.001", paste0("p = ", round(pv, 3)))

ord_plot(pcoa_ang_bray, color = "Gilliamella_PA", size = 4, alpha = 0.85) +
  scale_color_manual(values = c("Present" = "#f5a623", "Absent" = "#9E9E9E"),
                     name = "Gilliamella") +
  annotate("text", x = -Inf, y = Inf,
           label = paste0("R² = ", r2, ", ", plab),
           hjust = -0.1, vjust = 1.5, size = 3.5) +
  theme_classic() +
  ggtitle(expression(italic("Tetragonisca angustula"))) +
  coord_fixed()
```

---

# 7. PERMANOVA

## 7.1 All species

```{r permanova-all}
otu_all  <- as.data.frame(t(otu_table(physeq_clr)))
meta_all <- data.frame(sample_data(physeq_clr), stringsAsFactors = FALSE)

permanova_all <- adonis2(
  otu_all ~ Organism, data = meta_all,
  method = "euclidean", permutations = 999, by = "margin"
)
print(permanova_all)
```

## 7.2 Per species (by colony)

```{r permanova-per-species}
otu_angustula    <- as.data.frame(t(otu_table(physeq_angustula)))
otu_orizabaensis <- as.data.frame(t(otu_table(physeq_orizabaensis)))
meta_angustula    <- data.frame(sample_data(physeq_angustula),    stringsAsFactors = FALSE)
meta_orizabaensis <- data.frame(sample_data(physeq_orizabaensis), stringsAsFactors = FALSE)

permanova_angustula <- adonis2(
  otu_angustula ~ ColonyName, data = meta_angustula,
  method = "euclidean", permutations = 999, by = "margin"
)

permanova_orizabaensis <- adonis2(
  otu_orizabaensis ~ ColonyName, data = meta_orizabaensis,
  method = "euclidean", permutations = 999, by = "margin"
)

cat("=== Tetragonisca angustula ===\n"); print(permanova_angustula)
cat("\n=== Partamona orizabaensis ===\n"); print(permanova_orizabaensis)
```

## 7.3 Betadisper (homogeneity of dispersion)

```{r betadisper}
dist_angustula    <- dist(otu_angustula,    method = "euclidean")
dist_orizabaensis <- dist(otu_orizabaensis, method = "euclidean")

cat("=== Tetragonisca angustula ===\n")
permutest(betadisper(dist_angustula,    meta_angustula$ColonyName))

cat("\n=== Partamona orizabaensis ===\n")
permutest(betadisper(dist_orizabaensis, meta_orizabaensis$ColonyName))
```

---

# 8. Symbiont relative abundance

```{r symbiont-relabun}
ps_rel <- transform_sample_counts(physeq_clean, function(x) x / sum(x))

# Snodgrassella
ps_snod <- ps_rel %>%
  subset_taxa(Genus == "Snodgrassella") %>%
  tax_glom(taxrank = "Genus")

df_snod <- as.data.frame(otu_table(ps_snod)) %>%
  rownames_to_column("Taxa") %>%
  pivot_longer(-Taxa, names_to = "SampleID", values_to = "Abundance") %>%
  mutate(Taxa = "Snodgrassella")

# Gilliamella
ps_gill <- ps_rel %>%
  subset_taxa(Genus == "Gilliamella") %>%
  tax_glom(taxrank = "Genus")

df_gill <- as.data.frame(otu_table(ps_gill)) %>%
  rownames_to_column("Taxa") %>%
  pivot_longer(-Taxa, names_to = "SampleID", values_to = "Abundance") %>%
  mutate(Taxa = "Gilliamella")

df_symbionts <- bind_rows(df_snod, df_gill) %>%
  left_join(metadata_jb, by = "SampleID") %>%
  filter(!is.na(Abundance), !is.na(Organism))
```

```{r symbiont-mirror-histogram}
ggplot(df_symbionts, aes(x = Abundance, fill = Taxa)) +
  geom_histogram(
    data = filter(df_symbionts, Taxa == "Snodgrassella"),
    aes(y = ..count..), bins = 30, alpha = 0.7
  ) +
  geom_histogram(
    data = filter(df_symbionts, Taxa == "Gilliamella"),
    aes(y = -..count..), bins = 30, alpha = 0.7
  ) +
  facet_wrap(~ Organism) +
  scale_fill_manual(values = c("Gilliamella" = "#2b9dce", "Snodgrassella" = "#e54b4b")) +
  scale_y_continuous(labels = abs) +
  labs(x = "Relative abundance", y = "Number of bees") +
  coord_flip() +
  theme_minimal()
```

## 8.2 Fig S3 — Mirror histogram (log scale, P/A threshold)

```{r symbiont-mirror-log, fig.width=18, fig.height=5}
# Small offset so zero-abundance samples render on log scale
df_log <- df_symbionts %>%
  mutate(Abun_plot = pmax(Abundance, 1e-5))

ggplot(df_log, aes(x = Abun_plot, fill = Taxa)) +
  geom_histogram(
    data = filter(df_log, Taxa == "Snodgrassella"),
    aes(y = ..count..), bins = 25, alpha = 0.7
  ) +
  geom_histogram(
    data = filter(df_log, Taxa == "Gilliamella"),
    aes(y = -..count..), bins = 25, alpha = 0.7
  ) +
  geom_vline(xintercept = 0.001, linetype = "dashed",
             color = "black", linewidth = 0.6) +
  facet_wrap(~ Organism, nrow = 1,
             labeller = as_labeller(function(x) gsub("_", " ", x))) +
  scale_x_log10(
    breaks = c(1e-4, 1e-3, 1e-2, 1e-1, 1),
    labels = c("0.0001", "0.001", "0.01", "0.1", "1")
  ) +
  scale_fill_manual(values = c("Gilliamella"   = "#2b9dce",
                               "Snodgrassella" = "#e54b4b"),
                   labels = c("Gilliamella"   = expression(italic("Gilliamella")),
                               "Snodgrassella" = expression(italic("Snodgrassella")))) +
  scale_y_continuous(labels = abs) +
  labs(x = "Relative abundance (log scale)", y = "Number of bees",
       fill = NULL,
       caption = "Dashed line = 0.001 cutoff (presence/absence threshold).") +
  coord_flip() +
  theme_minimal() +
  theme(strip.text    = element_text(face = "italic"),
        plot.caption  = element_text(size = 8, color = "grey40"),
        panel.spacing = unit(1.5, "lines"))

ggsave("figures/raw/FigS3_symbiont_mirror_log.pdf", width = 18, height = 5)
```

---

# 9. Shannon diversity curves (Fig. S2)

Rarefaction of Shannon diversity by number of bees pooled.

```{r shannon-rarefaction-fn}
shannon_rarefaction <- function(asv_mat, iterations = 100) {
  n <- ncol(asv_mat)
  results <- matrix(NA_real_, nrow = n, ncol = iterations)
  for (i in seq_len(n)) {
    for (j in seq_len(iterations)) {
      sel <- sample(seq_len(n), i)
      sub <- if (i == 1) asv_mat[, sel, drop = FALSE] else rowSums(asv_mat[, sel, drop = FALSE])
      results[i, j] <- diversity(sub, index = "shannon")
    }
  }
  data.frame(
    n_samples     = seq_len(n),
    avg_diversity = rowMeans(results, na.rm = TRUE),
    std_err       = apply(results, 1, sd, na.rm = TRUE) / sqrt(iterations)
  )
}
```

```{r shannon-curves, fig.width=12, fig.height=8}
# ASV matrix (ASVs x samples) — forced to plain base R matrix
otu_raw  <- otu_table(physeq_clean)
otu_vals <- if (taxa_are_rows(physeq_clean)) slot(otu_raw, ".Data") else t(slot(otu_raw, ".Data"))
asv_mat  <- matrix(as.numeric(otu_vals), nrow = nrow(otu_vals), dimnames = dimnames(otu_vals))

meta_clean <- data.frame(sample_data(physeq_clean), stringsAsFactors = FALSE)
meta_clean$SampleID <- rownames(meta_clean)

make_rc_plot <- function(organism_name, color) {
  meta_sp <- filter(meta_clean, Organism == organism_name)
  mat_sp  <- asv_mat[, meta_sp$SampleID, drop = FALSE]
  mat_sp  <- mat_sp[rowSums(mat_sp) > 0, , drop = FALSE]
  res     <- shannon_rarefaction(mat_sp)

  ggplot(res, aes(n_samples, avg_diversity)) +
    geom_line(color = color) +
    geom_errorbar(aes(ymin = avg_diversity - std_err,
                      ymax = avg_diversity + std_err), width = 0.25) +
    labs(title = paste0("*", gsub("_", " ", organism_name), "*"),
         x = "Number of bees", y = "Average Shannon diversity") +
    ylim(1, 4) +
    theme_minimal() +
    theme(plot.title = element_markdown())
}

Rc_po <- make_rc_plot("Partamona_orizabaensis",        sp_colors_org[["Partamona_orizabaensis"]])
Rc_ss <- make_rc_plot("Scaptotrigona_subobscuripennis",sp_colors_org[["Scaptotrigona_subobscuripennis"]])
Rc_ta <- make_rc_plot("Tetragonisca_angustula",        sp_colors_org[["Tetragonisca_angustula"]])
Rc_tz <- make_rc_plot("Tetragona_ziegleri",           sp_colors_org[["Tetragona_ziegleri"]])
Rc_tf <- make_rc_plot("Trigona_fulviventris",          sp_colors_org[["Trigona_fulviventris"]])

Fig_S2 <- (Rc_po | Rc_ss | Rc_ta) / (Rc_tz | Rc_tf | plot_spacer())
Fig_S2
```

---

# 10. Collection map — Costa Rica

```{r costa-rica-map, fig.width=9, fig.height=7, message=FALSE, warning=FALSE}
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)
library(elevatr)
library(terra)
library(tidyterra)

# Reuse master palette (defined in species-colors chunk); map uses space-separated names
sp_colors <- setNames(sp_colors_org, gsub("_", " ", names(sp_colors_org)))

# Collection data (Table 2 — Joyce et al.)
df_map <- read.delim(paste0(path_raw, "CostaRica_metadata.tsv"),
                     stringsAsFactors = FALSE)
df_map <- df_map %>%
  mutate(
    Binomial  = paste(Genus, Species),
    Latitude  = as.numeric(Latitude),
    Longitude = as.numeric(Longitude)
  )

# Legend labels: "Genus species (n=samples, col=colonies)"
sp_summary <- df_map %>%
  group_by(Binomial) %>%
  summarise(
    n_samples  = n(),
    n_colonies = n_distinct(ColonyID),
    .groups    = "drop"
  ) %>%
  arrange(Binomial) %>%
  mutate(
    Species_label = paste0(Binomial, " (n=", n_samples, ", col=", n_colonies, ")")
  )

legend_colors <- setNames(sp_colors[sp_summary$Binomial], sp_summary$Species_label)

# One point per colony
df_unique <- df_map %>%
  group_by(ColonyID, Binomial) %>%
  summarise(
    Latitude  = first(Latitude),
    Longitude = first(Longitude),
    .groups   = "drop"
  ) %>%
  left_join(sp_summary %>% select(Binomial, Species_label), by = "Binomial") %>%
  mutate(Species_label = factor(Species_label, levels = sp_summary$Species_label))

# Basemap layers
costa_rica    <- ne_countries(country = "Costa Rica",              scale = "medium", returnclass = "sf")
neighbors     <- ne_countries(country = c("Nicaragua", "Panama"),  scale = "medium", returnclass = "sf")
provincias_sf <- ne_states(country = "Costa Rica", returnclass = "sf")

# Elevation raster — z=8 for better detail at zoomed extent
dem_terra <- elevatr::get_elev_raster(
  locations = costa_rica, z = 8, clip = "locations"
) %>%
  terra::rast() %>%
  terra::mask(terra::vect(costa_rica))

# Map
p <- ggplot() +

  # Neighboring countries
  geom_sf(data = neighbors,  fill = "grey90", color = "grey70", linewidth = 0.3) +

  # Costa Rica with elevation
  geom_sf(data = costa_rica, fill = "grey80", color = "grey50", linewidth = 1.0) +
  geom_spatraster(data = dem_terra) +
  scale_fill_gradientn(
    colours  = c("grey78", "grey60", "grey40", "grey18"),
    na.value = NA,
    name     = "Elevation (m asl)"
  ) +

  # Collection points — slight jitter separates species at same site
  geom_jitter(
    data   = df_unique,
    aes(x = Longitude, y = Latitude, color = Species_label),
    size   = 5.5,
    width  = 0.03,
    height = 0.03,
    alpha  = 0.7
  ) +

  # Scale bar
  annotation_scale(
    location   = "bl",
    width_hint = 0.15,
    height     = unit(0.15, "cm"),
    fill       = c("grey60", "white"),
    col        = "grey40"
  ) +

  # North arrow
  annotation_north_arrow(
    location = "tr",
    height   = unit(1, "cm"),
    width    = unit(1, "cm"),
    style    = north_arrow_fancy_orienteering(
      fill     = c("grey60", "white"),
      line_col = "grey40",
      text_col = "grey40"
    )
  ) +

  # Species colors — custom palette matching rest of document
  scale_color_manual(values = legend_colors, name = NULL) +

  # Extent — verified to cover all 9 colonies with margin
  coord_sf(xlim = c(-86.1, -84.4), ylim = c(9.95, 11.2), expand = FALSE) +

  labs(
    title   = "Stingless bee collection localities — Costa Rica",
    x       = NULL,
    y       = NULL,
    caption = "Data: Joyce (2018)"
  ) +

  theme_bw(base_size = 11) +
  theme(
    legend.position  = "right",
    legend.key.size  = unit(0.5, "cm"),
    legend.text      = element_text(size = 8, face = "italic"),
    plot.title       = element_text(face = "bold"),
    panel.grid.major = element_line(color = "grey95")
  )

p
```

---

```{r session-info, echo=FALSE}
sessionInfo()
```
